---
title: Bayesian linear models
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-float: true
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    html-math-method: katex
execute:
  keep-md: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: resources/references.bib
output_dir: "docs"
---


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)   #for data wrangling and plotting
library(DHARMa)      #for simulated residuals
library(performance) #for model diagnostics
library(see)         #for model diagnostics
library(brms)        #for Bayesian models
library(tidybayes)   #for exploring Bayesian PKPDmodels
library(rstan)       #for diagnostics plots
library(bayesplot)   #for diagnostic plots
library(patchwork)   #for arranging multiple plots
library(HDInterval)  #for HPD intervals
library(bayestestR)  #for ROPE
library(emmeans)     #for estimated marginal means
source("helperfunctions.R")
```

Many biologists and ecologists get a little twitchy and nervous around
mathematical and statistical formulae and nomenclature. Whilst it is
possible to perform basic statistics without too much regard for the
actual equation (model) being employed, as the complexity of the
analysis increases, the need to understand the underlying model
becomes increasingly important. Moreover, model specification in BUGS
(the language used to program Bayesian modelling) aligns very closely
to the underlying formulae. Hence a good understanding of the
underlying model is vital to be able to create a sensible Bayesian
model. Consequently, I will always present the linear model formulae
along with the analysis. If you start to feel some form of disorder
starting to develop, you might like to run through the Tutorials and
Workshops twice (the first time ignoring the formulae). 

# A philosophical note

To introduce the philosophical and mathematical differences between
classical (frequentist) and Bayesian statistics, @Wade-2000 presented
a provocative yet compelling trend analysis of two hypothetical
populations. The temporal trend of one of the populations shows very
little variability from a very subtle linear decline. By contrast, the
second population appears to decline more dramatically, yet has
substantially more variability.

@Wade-2000 neatly illustrates the contrasting conclusions
(particularly with respect to interpreting probability) that would be
drawn by the frequentist and Bayesian approaches and in so doing
highlights how and why the Bayesian approach provides outcomes that
are more aligned with management requirements.

This tutorial will start by replicating the demonstration of @Wade-2000.

```{r}
#| label: Pops
#| eval: true
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| tidy: true
set.seed(1)
x <- 1:10
b <- -0.03
a <- 200
sigma <- 0.11
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -0.1
a <- 200
sigma <- 0.38
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x)
#summary(lm(y~x))
popB <- data.frame(y,x)
#plot(y~x)

set.seed(1)
x <- seq(1,10,l=100)
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popC <- data.frame(y,x)
#plot(y~x)

#par(mar=c(3,3,0,0),mfrow=c(1,3))
#plot(y~x,popA, ylim=c(0,250))
#abline(lm(y~x,popA))
#plot(y~x,popB, ylim=c(0,250))
#abline(lm(y~x,popB))
#plot(y~x,popC, ylim=c(0,250))
#abline(lm(y~x,popC))

##THIS IS BETTER
for (n in 1:1000) {
  set.seed(n)
  x <- 1:10
  b <- -0.1
  a <- 200
  sigma <- 0.38
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popA <- data.frame(y,x))
  
  set.seed(n)
  x <- 1:10
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popB <- data.frame(y,x))
  
  set.seed(n)
  x <- seq(1,10,l=100)
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  popC <- data.frame(y,x)
  
  aa<-popA.lm <- lm(y~x,popA)
  bb<-popB.lm <- lm(y~x,popB)
  cc<-popC.lm <- lm(y~x,popC)
  AA<-summary(popA.lm)$coef
  BB<-summary(popB.lm)$coef
  CC<-summary(popC.lm)$coef
  
  if(AA[2,4]<0.05 & BB[2,4]>0.05 & abs(BB[2,1]-CC[2,1])<1) break
}
```
::: {.columns}

:::: {.column width="32%"}
```{r}
#| label: PopA
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popA, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popA))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popA.lm <- lm(y~x,popA)
popA.sum<-summary(popA.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popA)`<br>
Slope: `r round(coef(popA.lm)[2],4)`<br>
_t_: `r round(popA.sum[2,3],4)`<br>
_p-value_: `r round(popA.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopB
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popB, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popB))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popB.lm <- lm(y~x,popB)
popB.sum<-summary(popB.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popB)`<br>
Slope: `r round(coef(popB.lm)[2],4)`<br>
_t_: `r round(popB.sum[2,3],4)`<br>
_p-value_: `r round(popB.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopC
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
par(mar=c(3,3,0,0))
plot(y~x,popC, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popC))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popC.lm <- lm(y~x,popC)
popC.sum<-summary(popC.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popC)`<br>
Slope: `r round(coef(popC.lm)[2],4)`<br>
_t_: `r round(popC.sum[2,3],4)`<br>
_p-value_: `r round(popC.sum[2,4],4)`<br>
</p>

::::

:::


From a traditional frequentist perspective, we would conclude that
there is a 'significant' relationship in Population A and C
($p<0.05$), yet not in Population B ($p>0.05$). Note, Population B and
C were both generated from the same random distribution, it is just
that Population C has a substantially higher number of observations.

The above illustrates a couple of things

- statistical significance does not necessarily translate into
  biological importance. The percentage of decline for Population A is
  `r p1 <- predict(popA.lm, data.frame( x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)` where as the percentage of decline
  for Population B is `r p1 <- predict(popB.lm, data.frame(x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)`. That is Population B is declining
  at nearly 10 times the rate of Population A. That sounds rather
  important, yet on the basis of the hypothesis test, we would dismiss
  the decline in Population B.

- that a _p_-value is just the probability of detecting an effect or
  relationship - what is the probability that the sample size is large
  enough to pick up a difference.

Let us now look at it from a Bayesian perspective. I will just provide
the posterior distributions (densities scaled to 0-1 so that they can
be plotted together) for the slope for each population.

```{r}
#| label: Pop_densities
#| eval: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| warning: false
#| message: false
#| results: hide
#| cache: true
## PopA
popA.brm <- brm(y ~ x, data = popA)
slopeA <- popA.brm |>
    as_draws_df() |>
    pull("b_x")
## PopB
popB.brm <- brm(y ~ x, data = popB)
slopeB <- popB.brm |>
    as_draws_df() |>
    pull("b_x")
## PopC
popC.brm <- brm(y ~ x, data = popC)
slopeC <- popC.brm |>
    as_draws_df() |>
    pull("b_x")

ggplot() +
    geom_density(data = NULL, aes(x = slopeA, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeB, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeC, y = ..scaled..)) +
    scale_y_continuous("Relative probability", limits = c(0, 1), expand = c(0, 0)) +
    scale_x_continuous("Regression slope") +
  theme_classic() +
  theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
```

Focusing on Populations A and B, we would conclude:

- the mean (plus or minus CI) slopes for Population A and B are 
  `r round(mean(slopeA),2)` (`r paste(round(quantile(slopeA,
  p=c(0.025,0.975)),2),collapse=",")`) and `r round(mean(slopeB),2)`
  (`r paste(round(quantile(slopeB,
  p=c(0.025,0.975)),2),collapse=",")`) respectively.

- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.


# References

