---
title: Bayesian linear models
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-float: true
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    html-math-method: katex
execute:
  keep-md: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: resources/references.bib
output_dir: "docs"
---


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)   #for data wrangling and plotting
library(DHARMa)      #for simulated residuals
library(performance) #for model diagnostics
library(see)         #for model diagnostics
library(brms)        #for Bayesian models
library(tidybayes)   #for exploring Bayesian PKPDmodels
library(rstan)       #for diagnostics plots
library(bayesplot)   #for diagnostic plots
library(patchwork)   #for arranging multiple plots
library(HDInterval)  #for HPD intervals
library(bayestestR)  #for ROPE
library(emmeans)     #for estimated marginal means
source("helperfunctions.R")
```

Many biologists and ecologists get a little twitchy and nervous around
mathematical and statistical formulae and nomenclature. Whilst it is
possible to perform basic statistics without too much regard for the
actual equation (model) being employed, as the complexity of the
analysis increases, the need to understand the underlying model
becomes increasingly important. Moreover, model specification in BUGS
(the language used to program Bayesian modelling) aligns very closely
to the underlying formulae. Hence a good understanding of the
underlying model is vital to be able to create a sensible Bayesian
model. Consequently, I will always present the linear model formulae
along with the analysis. If you start to feel some form of disorder
starting to develop, you might like to run through the Tutorials and
Workshops twice (the first time ignoring the formulae). 

# A philosophical note

To introduce the philosophical and mathematical differences between
classical (frequentist) and Bayesian statistics, @Wade-2000 presented
a provocative yet compelling trend analysis of two hypothetical
populations. The temporal trend of one of the populations shows very
little variability from a very subtle linear decline. By contrast, the
second population appears to decline more dramatically, yet has
substantially more variability.

@Wade-2000 neatly illustrates the contrasting conclusions
(particularly with respect to interpreting probability) that would be
drawn by the frequentist and Bayesian approaches and in so doing
highlights how and why the Bayesian approach provides outcomes that
are more aligned with management requirements.

This tutorial will start by replicating the demonstration of @Wade-2000.

```{r}
#| label: Pops
#| eval: true
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| tidy: true
set.seed(1)
x <- 1:10
b <- -0.03
a <- 200
sigma <- 0.11
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -0.1
a <- 200
sigma <- 0.38
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x)
#summary(lm(y~x))
popB <- data.frame(y,x)
#plot(y~x)

set.seed(1)
x <- seq(1,10,l=100)
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popC <- data.frame(y,x)
#plot(y~x)

#par(mar=c(3,3,0,0),mfrow=c(1,3))
#plot(y~x,popA, ylim=c(0,250))
#abline(lm(y~x,popA))
#plot(y~x,popB, ylim=c(0,250))
#abline(lm(y~x,popB))
#plot(y~x,popC, ylim=c(0,250))
#abline(lm(y~x,popC))

##THIS IS BETTER
for (n in 1:1000) {
  set.seed(n)
  x <- 1:10
  b <- -0.1
  a <- 200
  sigma <- 0.38
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popA <- data.frame(y,x))
  
  set.seed(n)
  x <- 1:10
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popB <- data.frame(y,x))
  
  set.seed(n)
  x <- seq(1,10,l=100)
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  popC <- data.frame(y,x)
  
  aa<-popA.lm <- lm(y~x,popA)
  bb<-popB.lm <- lm(y~x,popB)
  cc<-popC.lm <- lm(y~x,popC)
  AA<-summary(popA.lm)$coef
  BB<-summary(popB.lm)$coef
  CC<-summary(popC.lm)$coef
  
  if(AA[2,4]<0.05 & BB[2,4]>0.05 & abs(BB[2,1]-CC[2,1])<1) break
}
```
::: {.columns}

:::: {.column width="32%"}
```{r}
#| label: PopA
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popA, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popA))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popA.lm <- lm(y~x,popA)
popA.sum<-summary(popA.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popA)`<br>
Slope: `r round(coef(popA.lm)[2],4)`<br>
_t_: `r round(popA.sum[2,3],4)`<br>
_p-value_: `r round(popA.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopB
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popB, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popB))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popB.lm <- lm(y~x,popB)
popB.sum<-summary(popB.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popB)`<br>
Slope: `r round(coef(popB.lm)[2],4)`<br>
_t_: `r round(popB.sum[2,3],4)`<br>
_p-value_: `r round(popB.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopC
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
par(mar=c(3,3,0,0))
plot(y~x,popC, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popC))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popC.lm <- lm(y~x,popC)
popC.sum<-summary(popC.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popC)`<br>
Slope: `r round(coef(popC.lm)[2],4)`<br>
_t_: `r round(popC.sum[2,3],4)`<br>
_p-value_: `r round(popC.sum[2,4],4)`<br>
</p>

::::

:::


From a traditional frequentist perspective, we would conclude that
there is a 'significant' relationship in Population A and C
($p<0.05$), yet not in Population B ($p>0.05$). Note, Population B and
C were both generated from the same random distribution, it is just
that Population C has a substantially higher number of observations.

The above illustrates a couple of things

- statistical significance does not necessarily translate into
  biological importance. The percentage of decline for Population A is
  `r p1 <- predict(popA.lm, data.frame( x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)` where as the percentage of decline
  for Population B is `r p1 <- predict(popB.lm, data.frame(x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)`. That is Population B is declining
  at nearly 10 times the rate of Population A. That sounds rather
  important, yet on the basis of the hypothesis test, we would dismiss
  the decline in Population B.

- that a _p_-value is just the probability of detecting an effect or
  relationship - what is the probability that the sample size is large
  enough to pick up a difference.

Let us now look at it from a Bayesian perspective. I will just provide
the posterior distributions (densities scaled to 0-1 so that they can
be plotted together) for the slope for each population.

```{r}
#| label: Pop_densities
#| eval: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| warning: false
#| message: false
#| results: hide
#| cache: true
## PopA
popA.brm <- brm(y ~ x, data = popA)
slopeA <- popA.brm |>
    as_draws_df() |>
    pull("b_x")
## PopB
popB.brm <- brm(y ~ x, data = popB)
slopeB <- popB.brm |>
    as_draws_df() |>
    pull("b_x")
## PopC
popC.brm <- brm(y ~ x, data = popC)
slopeC <- popC.brm |>
    as_draws_df() |>
    pull("b_x")

ggplot() +
    geom_density(data = NULL, aes(x = slopeA, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeB, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeC, y = ..scaled..)) +
    scale_y_continuous("Relative probability", limits = c(0, 1), expand = c(0, 0)) +
    scale_x_continuous("Regression slope") +
  theme_classic() +
  theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
```

Focusing on Populations A and B, we would conclude:

- the mean (plus or minus CI) slopes for Population A and B are 
  `r round(mean(slopeA),2)` (`r paste(round(quantile(slopeA,
  p=c(0.025,0.975)),2),collapse=",")`) and `r round(mean(slopeB),2)`
  (`r paste(round(quantile(slopeB,
  p=c(0.025,0.975)),2),collapse=",")`) respectively.

- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.
- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.

# Review of (generalised) linear models

I would highly recommend reviewing the information in [the tutorial on
generalised linear models](30_glm.html), particularly the sections
describe linear models, assumption checking and generalised linear
models (GLM). Whilst there are philosophical differences between
frequentist and Bayesian statistics that have implications for how
models are fit and interpreted, model choice and assumption checking
principles are common between the two approaches. Hence, many of these
topics will be assumed, and not fully described in the current
tutorial.

Recall from [the tutorial on generalised linear models](30_glm.html)
that simple linear regression is a linear modelling process that
models a single response variable against one or more predictors with
a linear combination of coefficients and (in the case of a Gaussian
model) can be expressed as:

::: {.columns}

:::: {.column width="48%"}
$$y_i = \beta_0+ \beta_1 x_i+\epsilon_i \hspace{1cm}\epsilon\sim{}N(0,\sigma^2)$$

where:

- $y_i$ is the response value for each of the $i$ observations
- $\beta_0$ is the y-intercept (value of $y$ when $x=0$)
- $\beta_1$ is the slope (rate of chance in $y$ per unit chance in
  $x$)
- $x_i$ is the predictor value for each of the $i$ observations
- $\epsilon_i$ is the residual value of each of the $i$ observations.
  A residual is the difference between the observed value and the
  value expected by the model.

- $\epsilon\sim{}N(0,\sigma^2)$ indicates that the residuals are
  normally distributed with a constant amount of variance

::::

:::: {.column width="48%"}

```{r}
#| label: residuals,
#| eval: true
#| echo: false
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
x=1:10
y=2+5*x
y1=2+5*x + rnorm(10,0,4) 
lm1 <- lm(y1 ~ x)
cc <- coef(lm1)
r <- resid(lm1)
#print(r)
ff <- fitted(lm1)
#print(ff)
g2=ggplot(data=NULL, aes(y=y1, x=x)) + 
    geom_segment(data = NULL, aes(y = ff, x = x, yend = y1, xend = x), colour = "red") +
    geom_point() +
    geom_abline(slope = cc[2], intercept = cc[1], colour = "blue") +
    #annotate(geom='text', y=40, x=2, label=expression(y[i]==2+5*x[i]+epsilon[i]), parse=TRUE, hjust=0) +
    annotate(geom='text', y=30, x=2.5, label=expression(residual~(epsilon[i])), parse=TRUE, hjust=1) +
    annotate(geom='text', y=15, x=5, label=expression(slope~(beta[1])), parse=TRUE, hjust=0) +
    annotate(geom='text', y=-5, x=1, label=expression(y-intercept~(beta[0])), parse=TRUE, hjust=0) +
    geom_curve(aes(x = 2, y = 27, xend = 3.8, yend = 23), 
             curvature = 0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 4.7, y = 15, xend = 3.7, yend = 18), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 0.7, y = -5, xend = -0.5, yend = -0.5), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    scale_y_continuous('y', limits = c(-10, 50)) +
    xlim(-2, 10) +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
g2
```
::::

:::

The above can be re-expressed and generalised as:
$$
y_i\sim{}Dist(\mu_i, ...)\\
g(\mu_i) = \beta_0+ \beta_1 x_i
$$

where:

- $Dist$ represents a distribution from the exponential family (such
  as Gaussian, Poisson, Binomial, etc)
- $...$ represents additional parameters relevant to the nominated
  distribution (such as $\sigma^2$: Gaussian, $n$: Binomial and
  $\phi$: Negative Binomial, etc)
- $g()$ represents the link function (e.g. log: Poisson, logit:
  Binomial, etc)

The reliability of any model depends on the degree to which the data
adheres to the model assumptions. Hence, as with frequentist models,
exploratory data analysis (EDA) is a vital component of Bayesian
modelling and since the model structures are similar between
frequentist and Bayesian approaches, so too is EDA.

# Bayesian (generalised) linear models

For the purpose of introduction, we will start by exploring a Gaussian
model with a very simple fabricated data set representing the
relationship between a response ($y$) and a continuous predictor ($x =
[1,2,3,4,5,6,7,8,9,10]$. The fabricated data set will comprise 10
observations each drawn from normal distributions with a set standard
deviation of 4. The means of the 10 populations will be determined by
the following equation: 

$$ \mu_i = 2 + 5\times x_i $$

Let generate these data.

```{r}
#| label: sim1
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
dat <- data.frame(x = 1:10) |>
    mutate(y = round(rnorm(n = 10, mean = 2 + (5 * x), sd = 4), digits = 2))
dat
```



The model will will be fitting will be:

$$
\begin{align}
y_i&\sim{}N(\mu_i, \sigma^2)\\
\mu_i &= \beta_0+ \beta_1 x_i
\end{align}
$$

The parameters that we are going to attempt to estimate are the
y-intercept ($\beta_0$), the slope ($\beta_1$) and the underlying
variance ($\sigma^2$). Recall (from tutorials on [statistical
philosophies](21_statistical_philosophies.html) and
[estimation](22_estimation.html))) that Bayesian models calculate
**posterior** predictions ($P(H|D)$) from likelihood ($P(D|H)$) and
**prior** expectations ($P(H)$). Therefore, in preparation for fitting
a Bayesian model, we must consider what our prior expectations are for
_all_ parameters.

The individual responses ($y_i$, observed yields) are each expected to
have been **independently** drawn from normal (**Gaussian**)
distributions ($\mathcal{N}$). These distributions represent all the
possible values of $y$ we could have obtained at the specific ($i^{th}$)
level of $x$. Hence the $i^{th}$ $y$ observation is expected
to have been drawn from a normal distribution with a mean of $\mu_i$.

Although each distribution is expected to come from populations that
differ in their means, we assume that all of these distributions have
the **same variance** ($\sigma^2$).



# References

