---
title: Bayesian linear models
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-float: true
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    html-math-method: katex
execute:
  keep-md: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: resources/references.bib
output_dir: "docs"
---

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
options(tinytex.engine = 'xelatex')
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)   #for data wrangling and plotting
library(DHARMa)      #for simulated residuals
library(performance) #for model diagnostics
library(see)         #for model diagnostics
library(brms)        #for Bayesian models
library(tidybayes)   #for exploring Bayesian PKPDmodels
library(rstan)       #for diagnostics plots
library(bayesplot)   #for diagnostic plots
library(patchwork)   #for arranging multiple plots
library(HDInterval)  #for HPD intervals
library(bayestestR)  #for ROPE
library(emmeans)     #for estimated marginal means
source("helperfunctions.R")
```

Many biologists and ecologists get a little twitchy and nervous around
mathematical and statistical formulae and nomenclature. Whilst it is
possible to perform basic statistics without too much regard for the
actual equation (model) being employed, as the complexity of the
analysis increases, the need to understand the underlying model
becomes increasingly important. Moreover, model specification in BUGS
(the language used to program Bayesian modelling) aligns very closely
to the underlying formulae. Hence a good understanding of the
underlying model is vital to be able to create a sensible Bayesian
model. Consequently, I will always present the linear model formulae
along with the analysis. If you start to feel some form of disorder
starting to develop, you might like to run through the Tutorials and
Workshops twice (the first time ignoring the formulae). 

# A philosophical note

To introduce the philosophical and mathematical differences between
classical (frequentist) and Bayesian statistics, @Wade-2000 presented
a provocative yet compelling trend analysis of two hypothetical
populations. The temporal trend of one of the populations shows very
little variability from a very subtle linear decline. By contrast, the
second population appears to decline more dramatically, yet has
substantially more variability.

@Wade-2000 neatly illustrates the contrasting conclusions
(particularly with respect to interpreting probability) that would be
drawn by the frequentist and Bayesian approaches and in so doing
highlights how and why the Bayesian approach provides outcomes that
are more aligned with management requirements.

This tutorial will start by replicating the demonstration of @Wade-2000.

```{r}
#| label: Pops
#| eval: true
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| tidy: true
set.seed(1)
x <- 1:10
b <- -0.03
a <- 200
sigma <- 0.11
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -0.1
a <- 200
sigma <- 0.38
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x)
#summary(lm(y~x))
popB <- data.frame(y,x)
#plot(y~x)

set.seed(1)
x <- seq(1,10,l=100)
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popC <- data.frame(y,x)
#plot(y~x)

#par(mar=c(3,3,0,0),mfrow=c(1,3))
#plot(y~x,popA, ylim=c(0,250))
#abline(lm(y~x,popA))
#plot(y~x,popB, ylim=c(0,250))
#abline(lm(y~x,popB))
#plot(y~x,popC, ylim=c(0,250))
#abline(lm(y~x,popC))

##THIS IS BETTER
for (n in 1:1000) {
  set.seed(n)
  x <- 1:10
  b <- -0.1
  a <- 200
  sigma <- 0.38
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popA <- data.frame(y,x))
  
  set.seed(n)
  x <- 1:10
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popB <- data.frame(y,x))
  
  set.seed(n)
  x <- seq(1,10,l=100)
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  popC <- data.frame(y,x)
  
  aa<-popA.lm <- lm(y~x,popA)
  bb<-popB.lm <- lm(y~x,popB)
  cc<-popC.lm <- lm(y~x,popC)
  AA<-summary(popA.lm)$coef
  BB<-summary(popB.lm)$coef
  CC<-summary(popC.lm)$coef
  
  if(AA[2,4]<0.05 & BB[2,4]>0.05 & abs(BB[2,1]-CC[2,1])<1) break
}
```
::: {.columns}

:::: {.column width="32%"}
```{r}
#| label: PopA
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popA, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popA))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popA.lm <- lm(y~x,popA)
popA.sum<-summary(popA.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popA)`<br>
Slope: `r round(coef(popA.lm)[2],4)`<br>
_t_: `r round(popA.sum[2,3],4)`<br>
_p-value_: `r round(popA.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopB
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popB, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popB))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popB.lm <- lm(y~x,popB)
popB.sum<-summary(popB.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popB)`<br>
Slope: `r round(coef(popB.lm)[2],4)`<br>
_t_: `r round(popB.sum[2,3],4)`<br>
_p-value_: `r round(popB.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopC
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
par(mar=c(3,3,0,0))
plot(y~x,popC, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popC))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popC.lm <- lm(y~x,popC)
popC.sum<-summary(popC.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popC)`<br>
Slope: `r round(coef(popC.lm)[2],4)`<br>
_t_: `r round(popC.sum[2,3],4)`<br>
_p-value_: `r round(popC.sum[2,4],4)`<br>
</p>

::::

:::


From a traditional frequentist perspective, we would conclude that
there is a 'significant' relationship in Population A and C
($p<0.05$), yet not in Population B ($p>0.05$). Note, Population B and
C were both generated from the same random distribution, it is just
that Population C has a substantially higher number of observations.

The above illustrates a couple of things

- statistical significance does not necessarily translate into
  biological importance. The percentage of decline for Population A is
  `r p1 <- predict(popA.lm, data.frame( x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)` where as the percentage of decline
  for Population B is `r p1 <- predict(popB.lm, data.frame(x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)`. That is Population B is declining
  at nearly 10 times the rate of Population A. That sounds rather
  important, yet on the basis of the hypothesis test, we would dismiss
  the decline in Population B.

- that a _p_-value is just the probability of detecting an effect or
  relationship - what is the probability that the sample size is large
  enough to pick up a difference.

Let us now look at it from a Bayesian perspective. I will just provide
the posterior distributions (densities scaled to 0-1 so that they can
be plotted together) for the slope for each population.

```{r}
#| label: Pop_densities
#| eval: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| warning: false
#| message: false
#| results: hide
#| cache: true
## PopA
popA.brm <- brm(y ~ x, data = popA)
slopeA <- popA.brm |>
    as_draws_df() |>
    pull("b_x")
## PopB
popB.brm <- brm(y ~ x, data = popB)
slopeB <- popB.brm |>
    as_draws_df() |>
    pull("b_x")
## PopC
popC.brm <- brm(y ~ x, data = popC)
slopeC <- popC.brm |>
    as_draws_df() |>
    pull("b_x")

ggplot() +
    geom_density(data = NULL, aes(x = slopeA, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeB, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeC, y = ..scaled..)) +
    scale_y_continuous("Relative probability", limits = c(0, 1), expand = c(0, 0)) +
    scale_x_continuous("Regression slope") +
  theme_classic() +
  theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
```

Focusing on Populations A and B, we would conclude:

- the mean (plus or minus CI) slopes for Population A and B are 
  `r round(mean(slopeA),2)` (`r paste(round(quantile(slopeA,
  p=c(0.025,0.975)),2),collapse=",")`) and `r round(mean(slopeB),2)`
  (`r paste(round(quantile(slopeB,
  p=c(0.025,0.975)),2),collapse=",")`) respectively.

- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.
- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.

# Review of (generalised) linear models

I would highly recommend reviewing the information in [the tutorial on
generalised linear models](30_glm.html), particularly the sections
describe linear models, assumption checking and generalised linear
models (GLM). Whilst there are philosophical differences between
frequentist and Bayesian statistics that have implications for how
models are fit and interpreted, model choice and assumption checking
principles are common between the two approaches. Hence, many of these
topics will be assumed, and not fully described in the current
tutorial.

Recall from [the tutorial on generalised linear models](30_glm.html)
that simple linear regression is a linear modelling process that
models a single response variable against one or more predictors with
a linear combination of coefficients and (in the case of a Gaussian
model) can be expressed as:

::: {.columns}

:::: {.column width="48%"}
$$y_i = \beta_0+ \beta_1 x_i+\epsilon_i \hspace{1cm}\epsilon\sim{}N(0,\sigma^2)$$

where:

- $y_i$ is the response value for each of the $i$ observations
- $\beta_0$ is the y-intercept (value of $y$ when $x=0$)
- $\beta_1$ is the slope (rate of chance in $y$ per unit chance in
  $x$)
- $x_i$ is the predictor value for each of the $i$ observations
- $\epsilon_i$ is the residual value of each of the $i$ observations.
  A residual is the difference between the observed value and the
  value expected by the model.

- $\epsilon\sim{}N(0,\sigma^2)$ indicates that the residuals are
  normally distributed with a constant amount of variance

::::

:::: {.column width="48%"}

```{r}
#| label: residuals,
#| eval: true
#| echo: false
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
x=1:10
y=2+5*x
y1=2+5*x + rnorm(10,0,4) 
lm1 <- lm(y1 ~ x)
cc <- coef(lm1)
r <- resid(lm1)
#print(r)
ff <- fitted(lm1)
#print(ff)
g2=ggplot(data=NULL, aes(y=y1, x=x)) + 
    geom_segment(data = NULL, aes(y = ff, x = x, yend = y1, xend = x), colour = "red") +
    geom_point() +
    geom_abline(slope = cc[2], intercept = cc[1], colour = "blue") +
    #annotate(geom='text', y=40, x=2, label=expression(y[i]==2+5*x[i]+epsilon[i]), parse=TRUE, hjust=0) +
    annotate(geom='text', y=30, x=2.5, label=expression(residual~(epsilon[i])), parse=TRUE, hjust=1) +
    annotate(geom='text', y=15, x=5, label=expression(slope~(beta[1])), parse=TRUE, hjust=0) +
    annotate(geom='text', y=-5, x=1, label=expression(y-intercept~(beta[0])), parse=TRUE, hjust=0) +
    geom_curve(aes(x = 2, y = 27, xend = 3.8, yend = 23), 
             curvature = 0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 4.7, y = 15, xend = 3.7, yend = 18), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 0.7, y = -5, xend = -0.5, yend = -0.5), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    scale_y_continuous('y', limits = c(-10, 50)) +
    xlim(-2, 10) +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
g2
```
::::

:::

The above can be re-expressed and generalised as:
$$
y_i\sim{}Dist(\mu_i, ...)\\
g(\mu_i) = \beta_0+ \beta_1 x_i
$$

where:

- $Dist$ represents a distribution from the exponential family (such
  as Gaussian, Poisson, Binomial, etc)
- $...$ represents additional parameters relevant to the nominated
  distribution (such as $\sigma^2$: Gaussian, $n$: Binomial and
  $\phi$: Negative Binomial, etc)
- $g()$ represents the link function (e.g. log: Poisson, logit:
  Binomial, etc)

The reliability of any model depends on the degree to which the data
adheres to the model assumptions. Hence, as with frequentist models,
exploratory data analysis (EDA) is a vital component of Bayesian
modelling and since the model structures are similar between
frequentist and Bayesian approaches, so too is EDA.

# Bayesian (generalised) linear models

For the purpose of introduction, we will start by exploring a Gaussian
model with a very simple fabricated data set representing the
relationship between a response ($y$) and a continuous predictor ($x =
[1,2,3,4,5,6,7,8,9,10]$. The fabricated data set will comprise 10
observations each drawn from normal distributions with a set standard
deviation of 4. The means of the 10 populations will be determined by
the following equation: 

$$ \mu_i = 2 + 5\times x_i $$

Let generate these data.

```{r}
#| label: sim1
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
dat <- data.frame(x = 1:10) |>
    mutate(y = round(rnorm(n = 10, mean = 2 + (5 * x), sd = 4), digits = 2))
dat
```



The model will will be fitting will be:

$$
\begin{align}
y_i&\sim{}N(\mu_i, \sigma^2)\\
\mu_i &= \beta_0+ \beta_1 x_i
\end{align}
$$

The parameters that we are going to attempt to estimate are the
y-intercept ($\beta_0$), the slope ($\beta_1$) and the underlying
variance ($\sigma^2$). Recall (from tutorials on [statistical
philosophies](21_statistical_philosophies.html) and
[estimation](22_estimation.html))) that Bayesian models calculate
**posterior** predictions ($P(H|D)$) from likelihood ($P(D|H)$) and
**prior** expectations ($P(H)$). Therefore, in preparation for fitting
a Bayesian model, we must consider what our prior expectations are for
_all_ parameters.

The individual responses ($y_i$, observed yields) are each expected to
have been **independently** drawn from normal (**Gaussian**)
distributions ($\mathcal{N}$). These distributions represent all the
possible values of $y$ we could have obtained at the specific ($i^{th}$)
level of $x$. Hence the $i^{th}$ $y$ observation is expected
to have been drawn from a normal distribution with a mean of $\mu_i$.

Although each distribution is expected to come from populations that
differ in their means, we assume that all of these distributions have
the **same variance** ($\sigma^2$).

## Priors

We need to supply priors for each of the parameters to be estimated
($\beta_0$, $\beta_1$ and $\sigma$). Whilst we want these priors to be
sufficiently vague as to not influence the outcomes of the analysis
(and thus be equivalent to the frequentist analysis), we do not want
the priors to be so vague (wide) that they permit the MCMC sampler to
drift off into parameter space that is both illogical as well as
numerically awkward.

Proffering sensible priors is one of the most difficult aspects of
performing Bayesian analyses. For instances where there are some
previous knowledge available and a desire to incorporate those data,
the difficulty is in how to ensure that the information is
incorporated correctly. However, for instances where there are no
previous relevant information and so a desire to have the posteriors
driven entirely by the new data, the difficulty is in how to define
priors that are both vague enough (not bias results in their
direction) and yet not so vague as to allow the MCMC sampler to drift
off into unsupported regions (and thus get stuck and yield spurious
estimates).

For early implementations of MCMC sampling routines (such as
Metropolis Hasting and Gibbs), it was fairly common to see very vague
priors being defined.  For example, the priors on effects, were
typically normal priors with mean of 0 and variance of `1e+06`
(1,00,000).  These are very vague priors.  Yet for some samplers
(e.g. NUTS), such vague priors can encourage poor behaviour of the
sampler - particularly if the posterior is complex. It is now
generally advised that priors should (where possible) be somewhat
**weakly informative** and to some extent, represent the bounds of
what are feasible and sensible estimates.

The degree to which priors __influence__ an outcome (whether by having
a pulling effect on the estimates or by encouraging the sampler to
drift off into unsupported regions of the posterior) is dependent on:

- the relative sparsity of the data - the larger the data, the less
  weight the priors have and thus less influence they exert.
- the complexity of the model (and thus posterior) - the more
  parameters, the more sensitive the sampler is to the priors.

The sampled posterior is the product of both the likelihood and the
prior - all of which are multidimensional.  For most applications, it
would be vertically impossible to define a sensible multidimensional
prior.  Hence, our only option is to define priors on individual
parameters (e.g. the intercept, slope(s), variance etc) and to hope
that if they are individually sensible, they will remain collectively
sensible.

So having (hopefully) impressed upon the notion that priors are an
important consideration, I will now attempt to synthesise some of the
approaches that can be employed to arrive at weakly informative priors
that have been gleaned from various sources.  Largely, this advice has
come from the following resources:

- https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations
- http://svmiller.com/blog/2021/02/thinking-about-your-priors-bayesian-analysis/

I will outline some of the current main recommendations before
summarising some approaches in a table.

- weakly informative priors should contain enough information so as to
  regularise (discourage unreasonable parameter estimates whilst
  allowing all reasonable estimates).
- for effects parameters on scaled (standardised) data, an argument
  could be made for a normal distribution with a standard deviation of
  1 (e.g. `normal(0,1)`), although some prefer a _t_ distribution with
  3 degrees of freedom and standard deviation of 1 (e.g.
  `student_t(3,0,1)`) - apparently a flatter _t_ is a more robust
  prior than a normal as an uninformative prior...
- for un-scaled data, the above priors can be scaled by using the
  standard deviation of the data as the prior standard deviation
  (e.g. `student_t(3,0,sd(y))`, or `sudent_t(3,0,sd(y)/sd(x))`)
- for priors of hierachical standard deviations, priors should
  encourage shrinkage towards 0 (particularly if the number of groups
  is small, since otherwise, the sampler will tend to be more
  responsive to "noise").

In this tutorial series, we will perform Bayesian analysis in the STAN
language via an R interface. Two popular interfaces that greatly
simplify the specification of Bayesian models are
[brms](https://paul-buerkner.github.io/brms/) and
[rstanarm](https://mc-stan.org/rstanarm/). We will exclusively focus
on the former as it is far more flexible.

| Family            | Parameter                            | brms                            | rstanarm                    |
|-------------------|--------------------------------------|---------------------------------|-----------------------------|
| Gaussian          | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | Sigma                                | `student_t(3,0,mad(y))`         | `exponential(1/sd(y))`      |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |
| Poisson           | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |
| Negative binomial | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | Shape                                | `gamma(0.01, 0.01)`             | `exponential(1/sd(y))`      |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |

: {.primary .bordered .sm .paramsTable}

Notes:

`brms`

https://github.com/paul-buerkner/brms/blob/c2b24475d727c8afd8bfc95947c18793b8ce2892/R/priors.R

1. In the above, for non-Gaussian families, `y` is first transformed
  according to the family link.  If the family link is `log`, then 0.1
  is first added to 0 values.
2. in `brms` the minimum standard deviation for the Intercept prior is
   `2.5`
3. in `brms` the minimum standard deviation for group-level priors is
   `10`.

`rstanarm`

http://mc-stan.org/rstanarm/articles/priors.html

1. in `rstanarm` priors on standard deviation and correlation
   associated with group-level effects are packaged up into a single
   prior (`decov` which is a decomposition of the variance and
   covariance matrix).

In my experience, I find that the above priors tend to be a little bit
too wide for many ecological applications and I often prefer to use
1.5 rather than 2.5 as the multiplier.

:::: {.callout-note}

In Bayesian models, centering of predictors offers huge numerical advantages.  So important is it to center that `brms` automatically centers any continuous predictors for you.  However, since the user has not necessarily centered the predictors, the user might misinterpret the outputs from a `brms` model.  Consequently, when fitting a model, `brms` also generates y-intercept values that are consistent with un-centered values and these are the estimates returned to the user.

Nevertheless, I would recommend that you should always explicitly
center continuous predictors to provide more meaningful
interpretations of the y-intercept. I would also highly recommend
standardising continuous predictors - this will not only help speed up
and stabilise the model, it will simplify the spefication of priors -
see the specific examples later in this tutorial.

::::

Based on the above, for our fabricated data, lets assign the following
priors:

- $\beta_0$: Normal prior centred at `r round(median(dat$y),2)` with a
  variance of `r round(mad(dat$y),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_1a
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> median() |> round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> mad() |> round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(dat$y)/mad(dat$x),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_1a
    #| echo: true
    #| eval: true
    #| cache: false
    mad(dat$y) / mad(dat$x) |> round(2)
    ```
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(mad(dat$y),2)`
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```

Note, again, when fitting models through either `rstanarm` or `brms`,
the priors assume that the predictor(s) have been centred and are to
be applied on the link scale. In this case the link scale is an
identity.



Similar logic can be applied for models that employ different
distributions. In the following sections, we will define numerous sets
of data (each of which represents different major forms of ecological
data) and see how we can set appropriate priors in each case. In
working through these example, it is worth reflecting on how much
simpler prior specification is if we use standardised predictors.


# Example data

This tutorial will blend theoretical discussions with actual
calculations and model fits. I believe that by bridging the divide
between theory and application, we all gain better understanding. The
applied components of this tutorial will be motivated by numerous
fabricated data sets. The advantage of simulated data over real data
is that with simulated data, we know the 'truth' and can therefore
gauge the accuracy of estimates.

The motivating examples are:

- **Example 1** - simulated samples drawn from a Gaussian (normal)
  distribution reminiscent of data collected on measurements (such as
  body mass)
- **Example 2** - simulated Gaussian samples drawn three different
  populations representing three different treatment levels (e.g. body
  masses of three different species)
- **Example 3** - simulated samples drawn from a Poisson distribution
  reminiscent of count data (such as number of individuals of a
  species within quadrats)
- **Example 4** - simulated samples drawn from a Negative Binomial
  distribution reminiscent of over-dispersed count data (such as
  number of individuals of a species that tends to aggregate in
  groups)
- **Example 5** - simulated samples drawn from a Bernoulli (binomial
  with $n = 1$) distribution reminiscent of binary data (such as the
  presence/absence of a species within sites)
- **Example 6** - simulated samples drawn from a Binomial distribution
  reminiscent of proportional data (such as counts of a particular
  taxa out of a total number of individuals)

::: {.panel-tabset}

## Example 1 (Gaussian data)

Lets formally simulate the data illustrated above. The underlying
process dictates that on average a one unit change in the predictor
(`x`) will be associated with a five unit change in response (`y`) and
when the predictor has a value of 0, the response will typically be 2.
Hence, the response (`y`) will be related to the predictor (`x`) via
the following:

$$
y = 2 + 5x
$$

This is a deterministic model, it has no uncertainty. In order to
simulate actual data, we need to add some random noise. We will assume
that the residuals are drawn from a Gaussian distribution with a mean
of zero and standard deviation of 4. The predictor will comprise of 10
uniformly distributed integer values between 1 and 10. We will round
the response to two decimal places.

For repeatability, a seed will be employed on the random number
generator. Note, the smaller the dataset, the less it is likely to
represent the underlying deterministic equation, so we should keep
this in mind when we look at how closely our estimated parameters
approximate the 'true' values. Hence, the seed has been chosen to
yield data that maintain a general trend that is consistent with the
defining parameters.

```{r}
#| label: sim
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
dat <- data.frame(x = 1:10) |>
    mutate(y = round(2 + 5*x + rnorm(n = 10, mean = 0, sd = 4), digits = 2))
dat
ggplot(data = dat) + 
geom_point(aes(y = y, x = x))
```

We will use these data in two ways. Firstly, to estimate the mean and
variance of the reponse (`y`) ignoring the predcitor (`x`) and
secondly to estimate the relationship between the reponse and
predictor.

For the former, we know that the mean and variance of the response
(`y`) can be calculated as:

$$
\begin{align}
\bar{y} =& \frac{1}{n}\sum^n_{i=1}y_i\\
var(y) =& \frac{1}{n}\sum^n_{i=1}(y-\bar{y})^2\\
sd(y) =& \sqrt{var(y)}
\end{align}
$$

```{r}
#| label: stats1
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
mean(dat$y)
var(dat$y)
sd(dat$y)
```

## Example 2 (categorical predictor)

As previously described, categorical predictors are transformed into
dummy codes prior to the fitting of the linear model. We will simulate
a small data set with a single categorical predictor comprising a
control and two treatment levels ('mediam', 'high'). To simplify
things we will assume a Gaussian distribution, however most of the
modelling steps would be the same regardless of the chosen
distribution.

The data will be drawn from three Gaussian distributions with a
standard deviation of 4 and means of 20, 15 and 10. We will draw a
total of 12 observations, four from each of the three populations.

```{r}
#| label: sim2
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(123)
beta_0 <- 20
beta <- c(-2, -10)
sigma <- 4
n <- 12
x <- gl(3, 4, 12, labels = c('control', 'medium', 'high'))
y <- (model.matrix(~x) %*% c(beta_0, beta)) + rnorm(12, 0, sigma)
dat2 <- data.frame(x = x, y = y)
dat2
ggplot(data = dat2) + 
geom_point(aes(y = y, x = x)) 
```

## Example 3 (Poisson data)

The Poisson distribution is only parameterized by a single parameter
($\lambda$) which represents both the mean and variance. Furthermore,
Poisson data can only be positive integers.  

Unlike simple trend between two Gaussian or uniform distributions,
modelling against a Poisson distribution alters the scale to
logarithms. This needs to be taken into account when we simulate the
data. The parameters that we used to simulate the underlying processes
need to either be on a logarithmic scale, or else converted to a
logarithmic scale prior to using them for generating the random data.

Moreover, for any model that involves a non-identity link function
(such as a logarithmic link function for Poisson models), 'slope' is
only constant on the scale of the link function. When it is back
transformed onto the natural scale (scale of the data), it takes on a
different meaning and interpretation.

We will chose $\beta_0$ to represent a value of 1 when `x=0`. As for
the 'effect' of the predictor on the response, lets say that for every
one unit increase in the predictor the response increases by 40% (on
the natural scale). Hence, on the log scale, the slope will be
$log(1.5)=$ `r log(1.4)`.


```{r}
#| label: sim3
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(123)
beta <- c(1, 1.40)
beta <- log(beta)
n <- 10
dat3 <- data.frame(x=seq(from = 1, to = 10, len = n)) |>
    mutate(y = rpois(n, lambda = exp(beta[1] + beta[2]*x)))
dat3
ggplot(data = dat3) + 
geom_point(aes(y = y, x = x)) 
```

## Example 4 (NB data)

In theory, count data should follow a Poisson distribution and
therefore have properties like mean equal to variance (e.g.
$\textnormal{Dispersion}=\frac{\sigma}{\mu}=1$). However as simple
linear models are low dimensional representations of a system, it is
often unlikely that such a simple model can capture all the
variability in the response (counts). For example, if we were
modelling the abundance of a species of intertidal snail within
quadrats in relation to water depth, it is highly likely that water
depth alone drives snail abundance. There are countless other
influences that the model has not accounted for. As a result, the
observed data might be more variable than a Poisson (of a particular
mean) would expect and in such cases, the model is over-dispersed
(more variance than expected).

Over dispersed models under-estimate the variability and thus
precision in estimates resulting in inflated confidence in outcomes
(elevated Type I errors).

There are numerous causes of over-dispersed count data (one of which
is eluded to above). These are:

- additional sources of variability not being accounted for in the
  model (see above)
- when the items being counted aggregate together. Although the
  underlying items may have been generated by a Poisson process, the
  items clump together. When the items are counted, they are more
  likely to be in either in relatively low or relatively high
  numbers - hence the data are more varied than would be expected from
  their overall mean.
- imperfect detection resulting in excessive zeros. Again the
  underlying items may have been generated by a Poisson process,
  however detecting and counting the items might not be completely
  straight forward (particularly for more cryptic items). Hence, the
  researcher may have recorded no individuals in a quadrat and yet
  there was one or more present, they were just not obvious and were
  not detected. That is, layered over the Poisson process is another
  process that determines the detectability. So while the Poisson
  might expect a certain proportion of zeros, the observed data might
  have a substantially higher proportion of zeros - and thus higher
  variance.

This example will generate data that is drawn from a negative binomial
distribution so as to broadly represent any one of the above causes.

We will chose $\beta_0$ to represent a value of 1 when `x=0`. As for
the 'effect' of the predictor on the response, lets say that for every
one unit increase in the predictor the response increases by 40% (on
the natural scale). Hence, on the log scale, the slope will be
$log(1.5)=$ `r log(1.4)`. Finally, the dispersion parameter (ratio of
variance to mean) will be 10.


```{r}
#| label: sim4
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
beta <- c(1, 1.40)
beta <- log(beta)
n <- 10
size <- 10
dat4 <- data.frame(x = seq(from = 1, to = 10, len = n)) |>
    mutate(
        mu = exp(beta[1] + beta[2] * x),
        y = rnbinom(n, size = size, mu =  mu)
    )
dat4
ggplot(data = dat4) + 
geom_point(aes(y = y, x = x)) 
```

## Example 5 (Binary data)

Binary data (presence/absence, dead/alive, yes/no, heads/tails, etc)
pose unique challenges for linear modeling. Linear regression,
designed for continuous outcomes, may not be directly applicable to
binary responses. The nature of binary data violates assumptions of
normality and homoscedasticity, which are fundamental to linear
regression. Furthermore, linear models may predict probabilities
outside the [0, 1] range, leading to unrealistic predictions.

This example will generate data that is drawn from a bernoulli
distribution so as to broadly represent presence/absence data.

We will chose $\beta_0$ to represent the odds of a value of 1 when
$x=0$ equal to $0.02$. This is equivalent to a probability of $y$
being zero when $x=0$ of $\frac{0.02}{1+0.02}=0.0196$. E.g., at low
$x$, the response is likely to be close to 0. For every one unit
increase in $x$, we will stipulate a 2 times increase in odds that
the expected response is equal to 1.


```{r}
#| label: sim5
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
beta <- c(0.02, 2)
beta <- log(beta)
n <- 10
dat5 <- data.frame(x = seq(from = 1, to = 10, len = n)) |>
    mutate(
        y = as.numeric(rbernoulli(n, p = plogis(beta[1] + beta[2] * x)))
    )
dat5
ggplot(data = dat5) + 
geom_point(aes(y = y, x = x)) 
```

## Example 6 (Binomial data)

Similar to binary data, proportional (binomial) data tend to violate
normality and homogeneity of variance (particularly as mean
proportions approach either 0% or 100%.

This example will generate data that is drawn from a binomial
distribution so as to broadly represent proportion data.

We will chose $\beta_0$ to represent the odds of a particular trial
(e.g. an individual) being of a particular type (e.g. species 1) when
$x=0$ and to equal to $0.02$. This is equivalent to a probability of
$y$ being of the focal type when $x=0$ of
$\frac{0.02}{1+0.02}=0.0196$. E.g., at low $x$, the the probability
that an individual is taxa 1 is likely to be close to 0. For every one
unit increase in $x$, we will stipulate a 2.5 times increase in odds
that the expected response is equal to 1.

For this example, we will also convert the counts into proportions
($y2$) by division with the number of trials ($5$).

```{r}
#| label: sim6
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(123)
beta <- c(0.02, 2.5)
beta <- log(beta)
n <- 10
trials <- 5
dat6 <- data.frame(x = seq(from = 1, to = 10, len = n)) |>
    mutate(
      count = as.numeric(rbinom(n, size = trials, prob = plogis(beta[1] + beta[2] * x))),
      total = trials,
      y = count/total
    )
dat6
ggplot(data = dat6) + 
geom_point(aes(y = y, x = x)) 
```


:::


# Exploratory data analysis

Statistical models utilize data and the inherent statistical
properties of distributions to discern patterns, relationships, and
trends, enabling the extraction of meaningful insights, predictions,
or inferences about the phenomena under investigation. To do so,
statistical models make assumptions about the likely distributions
from which the data were collected. Consequently, the reliability and
validity of any statistical model depend upon adherence to these
underlying assumptions.

Exploratory Data Analysis (EDA) and assumption checking therefore play
pivotal roles in the process of statistical analysis, offering
essential tools to glean insights, assess the reliability of
statistical methods, and ensure the validity of conclusions drawn from
data. EDA involves visually and statistically examining datasets to
understand their underlying patterns, distributions, and potential
outliers. These initial steps provides an intuitive understanding of
the data's structure and guides subsequent analyses. By scrutinizing
assumptions, such as normality, homoscedasticity, and independence,
researchers can identify potential limitations or violations that may
impact the accuracy and reliability of their findings.

Exploratory Data Analysis within the context of ecological statistical
models usually comprise a set of targetted graphical summaries. They
are not to be considered definitive diagnostics of the model
assumptions, but rather a first pass to assess the obvious violations
prior to the fitting of models. More definitive diagnostics can only
be achieved after a model has been fit. 

In addition to graphical summaries, there are numerous statistical
tests to help explore possible violations of various statistical
assumptions. These tests are less commonly used in ecology since they
are often more sensitive to deviations from ideal than are the models
that we are seeking to ensure. 

Simple classic regression models are often the easiest models to fit
and interpret and as such often represent a standard by which other
alternate models are gauged. As you will see later in this tutorial,
such models can actually be fit using closed form (exact solution)
matrix algebra that can be performed by hand. Nevertheless, and
perhaps as a result, they also impose some of the strictest
assumptions. Although these collective assumptions are specific to
gaussian models, they do provide a good introduction to model
assumptions in general, so we will use them to motivate the more wider
discussion.

Simple (gaussian) linear models (represented below) make the following
assumptions:

::: {.columns}

:::: {.column width="65%"}

```{tikz}
%| label: model1
%| eval: true
%| echo: false
%| cache: false
%| fig-width: 6
\usetikzlibrary{shapes,arrows,shadows,positioning,mindmap,backgrounds,decorations, calc,fit, decorations.pathreplacing,decorations.pathmorphing, shadings,shapes.geometric,patterns} 

\newcommand{\tikzmark}[1]{\tikz[baseline, remember picture] \coordinate (#1) {};}

\begin{tikzpicture}[remember picture]
\node (model) {$y_i=\underbrace{\beta_0 + \beta_1\times x_i}_\text{Linearity} + \varepsilon_i $};
\node [anchor = west, right = 2cm of model.east] (dist) {$\varepsilon_i\sim \underbrace{\mathcal{N}(0, \tikzmark{sigma}\sigma^2)}_\text{Normality}$};
\node [anchor = west, below = 1.25cm of model.east] (covnode) {$
\tikzmark{cov}\mathbf{V}=cov = \begin{pmatrix}
\tikzmark{covsigma}\sigma^2&0~&\cdots&\tikzmark{otherzero}0\\
0&\sigma^2&\cdots&\vdots\\
\vdots&\cdots&\sigma^2&\vdots\\
\tikzmark{zero}0&\cdots&\cdots&\tikzmark{othercovsigma}\sigma^2\\
\end{pmatrix} $};

\coordinate (sigmamark) at ($(othercovsigma.north) + (covnode.west|-covnode)$);
\path (sigmamark) +(1,0em) node[anchor = west](c) {Homogeneity of variance};
\draw[->] (sigmamark)++(.9em,0ex) to [bend right] (c.west) ;

\coordinate (zeromark) at ($(otherzero.north) + (covnode.west|-covnode)$);
\path (zeromark) +(1,-1em) node[anchor = west](Z) {\parbox{3cm}{Zero covariance\\ (=independence)}};
\draw[->] (zeromark)++(.9em,0ex) to [bend left] ($(Z.north west) + (0, -0.5em)$) ;
 
\coordinate (sigma2mark) at ($(sigma.north) + (dist.west|-dist)$);
\coordinate (covmark) at ($(cov.north) + (covnode.west|-covnode)$);

\draw[->,thick, black!50!white!50] ($(sigma2mark.east) +(1em, 0.5em)$) 
    .. controls ++(1cm,0.2cm) and ++(1.7cm,0.8em) .. ++(1cm,-1cm) 
    .. controls ++(-2cm,-0.1cm) and ++(4cm,-.0cm) ..  ++(-6cm,0cm)
    .. controls ++(-0.8cm, 0cm) and ++ (0.2cm, 1.3cm) .. ($(covmark.north) +(0.7em, 0.7em)$);
\end{tikzpicture}
```
 

::::

:::: {.column width="35%"}

```{r}
#| label: assumptions,
#| eval: true
#| echo: false
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
x=1:10
y=2+5*x
y1=2+5*x + rnorm(10,0,4) 
lm1 <- lm(y1 ~ x)
cc <- coef(lm1)
r <- resid(lm1)
#print(r)
ff <- fitted(lm1)
#print(ff)
boxx <- data.frame(x = x, middle = ff, lower = ff - 2.7, upper = ff + 2.7, ymin = ff -11, ymax = ff + 11)
g2=ggplot(data=NULL, aes(y=y1, x=x)) +
    geom_blank(aes(x = factor(x))) +
    ##geom_segment(data = NULL, aes(y = ff, x = x, yend = y1, xend = x), colour = "red") +
    geom_boxplot(data = boxx, aes(group = x, x = x, ymin = ymin, ymax = ymax, lower = lower, upper = upper, middle = middle), stat = "identity", inherit.aes = FALSE, color = "grey") +
    geom_point() +
    geom_abline(slope = cc[2], intercept = cc[1], colour = "blue") +
    scale_y_continuous('y') +
    theme_classic() #+
    #theme(axis.text = element_blank(), axis.ticks = element_blank())
g2
```

::::

:::


::: {.callout-note collapse="true"}
### Notes on the data depicted above
The data depicted above where generated using the following R codes:

```{r}
#| label: assumptions1,
#| eval: true
#| echo: true
#| warning: false
set.seed(234)
x <- 1:10
y <- 2 + (5*x) + rnorm(10,0,4)
```

The observations represent

- single observations drawn from 10 normal populations
- each population had a standard deviation of 4
- the mean of each population varied linearly according to the value of x ($2 + 5x$)
:::

- **normality**: the residuals (and thus observations) must be drawn
  from populations that are normal distribution. _The right hand figure
  underlays the ficticious normally distributed populations from which
  the observed values have been sampled_.
  
:::: {.indented}

::: {.callout-note collapse="true"}
### More information about assessing normality

Estimation and inference testing in linear regression assumes that the
response is normally distributed in each of the populations. In this
case, the populations are all possible measurements that could be
collected at each level of $x$ - hence there are 16 populations.
Typically however, we only collect a single observation from each
population (as is also the case here). How then can be evaluate
whether each of these populations are likely to have been normal?

::::: {.columns}

:::::: {.column width="45%"}

```{r}
#| label: assumptions1a
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 4
#| fig-height: 3
set.seed(1)
n <- 16
rep<-200
a <- 40  #intercept
b <- seq(0,15,l=16)  #slope
sigma2 <- 4  #residual variance (sd=2)
x <- gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
nc <- matrix(c(1,1,1,2,1,1,1,2),nrow=2,ncol=4,byrow=T)
lay <- layout(nc)
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
box(bty="l")
par(mar=c(3,3,0,0))
boxplot(y, axes=F, ann=FALSE)
```

::::::

:::::: {.column width="45%"}

```{r}
#| label: assumptions1b
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 4
#| fig-height: 3
set.seed(1)
n <- 16
rep<-200
a <- 40  #intercept
b <- seq(0,15,l=16)  #slope
sigma2 <- 3  #residual variance (sd=2)
x <- gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- (rnorm(n*rep, mean = 0, sd = sqrt(sigma2)))^{-1/3}  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
nc <- matrix(c(1,1,1,2,1,1,1,2),nrow=2,ncol=4,byrow=T)
lay <- layout(nc)
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
box(bty="l")
par(mar=c(3,3,0,0))
boxplot(y, axes=F, ann=FALSE)
```

::::::
:::::

For a given response, the population distributions should follow much
the same distribution shapes. Therefore provided the single samples
from each population are unbiased representations of those
populations, a boxplot of all observations should reflect the
population distributions.

The two figures above show the relationships between the individual
population distributions and the overall distribution. The left hand
figure shows a distribution drawn from single representatives of each
of the 16 populations. Since the 16 individual populations were
normally distributed, the distribution of the 16 observations is also
normal.

By contrast, the right hand figure shows 16 log-normally distributed
populations and the resulting distribution of 16 single observations
drawn from these populations. The overall boxplot mirrors each of the
individual population distributions.

Whilst traditionally, non-normal data would typically be
**normalised** via a scale transformation (such as a logarithmic
transformation), these days it is arguably more appropriate to attempt
to match the data to a more suitable distribution (see later in this
tutorial).

You may have noticed that we have only explored the distribution of
the response (y-axis). What about the distribution of the predictor
(independent, x-axis) variable, does it matter? The distribution
assumption applies to the residuals (which as purely in the direction
of the y-axis). Indeed technically, it is assumed that there is no
uncertainty associated with the predictor variable. They are assumed
to be set and thus there is no error associated with the values
observed. Whilst this might not always be reasonable, it is an
assumption.

Given that the predictor values are expected to be _set_ rather than
_measured_, we actually assume that they are **uniformly**
distributed. In practice, the exact distribution of predictor values
is not that important provided it is reasonably symmetrical and no
outliers (unusually small or large values) are created as a result of
the distribution.

As with exploring the distribution of the response variable, boxplots,
histograms and density plots can be useful means of exploring the
distribution of predictor variable(s). When such diagnostics reveal
distributional issues, scale transformations (such as logarithmic
transformations) are appropriate.

:::

::::



- **homogeneity of variance**: the residuals (and thus observations)
  must be drawn from populations that are equally varied. The model as
  shown only estimates a single variance ($\sigma^2$) parameter - it is
  assumed that this is a good overall representation of all underlying
  populations. _The right hand figure underlays the ficticious normally
  distributed and equally varied populations from which the
  observations have been sampled_.
  
  Moreover, since the expected values (obtained by solving the
  deterministic component of the model) and the variance must be
  estimated from the same data, they need to be independent (not
  related one another)

:::: {.indented}

::: {.callout-note collapse="true"}
### More information about assessing homogeneity of variance

Simple linear regression also assumes that each of the populations are
equally varied. Actually, it is the prospect of a relationship between
the mean and variance of y-values across x-values that is of the
greatest concern. Strictly the assumption is that the distribution of
y values at each x value are equally varied and that there is no
relationship between mean and variance.

However, as we only have a single y-value for each x-value, it is
difficult to directly determine whether the assumption of homogeneity
of variance is likely to have been violated (mean of one value is
meaningless and variability can't be assessed from a single value).
The figure below depicts the ideal (and almost never realistic)
situation in which (left hand figure) the populations are all equally
varied. The middle figure simulates drawing a single observation from
each of the populations. When the populations are equally varied, the
spread of observed values around the trend line is fairly even - that
is, there is no trend in the spread of values along the line.

If we then plot the residuals (difference between observed values and
those predicted by the trendline) against the predict values, there is
a definite lack of pattern. This lack of pattern is indicative of a
lack of issues with homogeneity of variance.

::::: {.columns}

:::::: {.column width="32%"}

```{r}
#| label: assumptions2a
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(1)
n <- 16
rep<-200
a <- 40  #intercept
b <- seq(0,15,l=16)  #slope
sigma2 <- 4  #residual variance (sd=2)
x <- gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
mtext("X",1,line=1, cex=1.5)
mtext("Y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2b
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(1)
n <- 16
rep<-1
a <- 40  #intercept
b <- 15  #slope
sigma2 <- 1000  #residual variance (sd=5)
x<-1:16

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
segments(x,y,x,fitted(lm(y~x)))
mtext("x",1,line=1, cex=1.5)
mtext("y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2c
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
res <- y-fitted(lm(y~x))
exp <- fitted(lm(y~x))
par(mar=c(3,3,0,0))
plot(exp,res, xlab="Predicted x", type="p", pch=16, axes=F)
mtext("Predicted x",1,line=1, cex=1.5)
mtext("Residuals",2,line=1, cex=1.5)
abline(h=0)
box(bty="l")
```

::::::
:::::

If we now contrast the above to a situation where the population
variance is related to the mean (unequal variance), we see that the
observations drawn from these populations are not evenly distributed
along the trendline (they get more spread out as the mean predicted
value increase). This pattern is emphasized in the residual plot which
displays a characteristic "wedge"-shape pattern.

::::: {.columns}

:::::: {.column width="32%"}

```{r}
#| label: assumptions2d
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(1)
n <- 16
rep<-200
a <- 40  #intercept
b <- seq(0,15,l=16)  #slope
sigma2 <- rep(seq(1,200,l=16),each=rep)  #residual variance (sd=2)
x <- gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
mtext("X",1,line=1, cex=1.5)
mtext("Y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2e
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(7)
n <- 16
rep<-1
a <- 40  #intercept
b <- 15  #slope
sigma2 <- seq(1,2000,l=16)  #residual variance (sd=5)
x<-1:16

eps <- x*10#+rnorm(n,0,(1:16)*7)
eps[(1:16 %%2)==1] <- -1*eps[(1:16 %%2)==1]
#rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~x) %*% c(a,b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
segments(x,y,x,fitted(lm(y~x)))
mtext("x",1,line=1, cex=1.5)
mtext("y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2f
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
res <- y-fitted(lm(y~x))
exp <- fitted(lm(y~x))
par(mar=c(3,3,0,0))
plot(exp,res, xlab="Predicted x", type="p", pch=16, axes=F)
mtext("Predicted x",1,line=1, cex=1.5)
mtext("Residuals",2,line=1, cex=1.5)
abline(h=0)
box(bty="l")
```

::::::
:::::

Hence looking at the spread of values around a trendline on a
scatterplot of $y$ against $x$ is a useful way of identifying gross
violations of homogeneity of variance. Residual plots provide an even
better diagnostic. The presence of a wedge shape is indicative that
the population mean and variance are related.
:::

::::

- **linearity**: the underlying relationships must be simple linear
  trends, since the line of best fit through the data (of which the
  slope is estimated) is linear. _The right hand figure depicts a
  linear trend through the underlying populations_.

:::: {.indented}

::: {.callout-note collapse="true"}
### More information about assessing linearity

It is important to disclose the meaning of the word "linear" in the
term "linear regression". Technically, it refers to a _linear_
combination of regression coefficients. For example, the following are
examples of linear models:

- $y_i = \beta_0 + \beta_1 x_i$
- $y_i = \beta_0 + \beta_1 x_i + \beta_2 z_i$
- $y_i = \beta_0 + \beta_1 x_i + \beta_2 x^2_i$

All the coefficients ($\beta_0$, $\beta_1$, $\beta_2$) are linear
terms. Note that the last of the above examples, is a linear model,
however it describes a non-linear trend. Contrast the above models
with the following non-linear model:

- $y_i = \beta_0 + x_i^{\beta_1}$

In that case, the coefficients are not linear combinations (one of
them is a power term).


That said, a simple linear regression usually fits a straight (linear)
line through the data. Therefore, prior to fitting such a model, it is
necessary to establish whether this really is the most sensible way of
describing the relationship. That is, does the relationship appear to
be linearly related or could some other non-linear function describe
the relationship better. Scatterplots and residual plots are useful
diagnostics.

To see how a residual plot could be useful, consider the following.
The first row of figures illustrate the residuals resulting from data
drawn from a linear trend. The residuals are effectively random noise.
By contrast, the second row show the residuals resulting from data
drawn from a non-normal relationship that have nevertheless been
modelled as a linear trend. There is still a clear pattern remaining
in the residuals.

::::: {.columns}

:::::: {.column width="32%"}

```{r}
#| label: assumptions2a
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2b
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2c
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
```

::::::
:::::

::::: {.columns}

:::::: {.column width="32%"}

```{r}
#| label: assumptions2a2
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(1)
n <- 16
rep<-200
a <- 40  #intercept
b <-c(50, 100)
sigma2 <- 4  #residual variance (sd=2)
x <- gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~poly(x, 2)) %*% c(a, b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
mtext("X",1,line=1, cex=1.5)
mtext("Y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2b2
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
set.seed(1)
n <- 16
rep<-1
a <- 40  #intercept
b <-c(50, 100)
sigma2 <- 4  #residual variance (sd=2)
x <- 1:16 #gl(n,rep,n*rep,lab=0:16)  #values of the year covariate

eps <- rnorm(n*rep, mean = 0, sd = sqrt(sigma2))  #residuals
y <- (model.matrix(~poly(x, 2)) %*% c(a, b))+eps
par(mar=c(3,3,0,0))
plot(y~x, las=1, axes=FALSE, ann=FALSE)
abline(lm(y~as.numeric(x)), col="grey40")
segments(x,y,x,fitted(lm(y~x)))
mtext("x",1,line=1, cex=1.5)
mtext("y",2,line=1, cex=1.5, las=2)
box(bty="l")
```

::::::

:::::: {.column width="32%"}

```{r}
#| label: assumptions2c2
#| eval: true
#| echo: false
#| warning: false
#| fig-width: 2.5
#| fig-height: 2
res <- y-fitted(lm(y~x))
exp <- fitted(lm(y~x))
par(mar=c(3,3,0,0))
plot(exp,res, xlab="Predicted x", type="p", pch=16, axes=F)
mtext("Predicted x",1,line=1, cex=1.5)
mtext("Residuals",2,line=1, cex=1.5)
abline(h=0)
box(bty="l")
```

::::::
:::::

The above might be an obvious and somewhat overly contrived example,
yet it does illustrate the point - that a pattern in the residuals
could point to a mis-specified model.

If non-linearity does exist (as in the second case above) , then
fitting a straight line through what is obviously not a straight
relationship is likely to poorly represent the true nature of the
relationship. There are numerous causes of non-linearity:

1. underlying distributional issues can result in non-linearity. For
   example, if we are assuming a gaussian distribution and the data
   are non-normal, often the relationships will appear non-linear.
   Addressing the distributional issues can therefore resolve the
   linearity issues
2. the underlying relationship might truly be non-linear in which case
   this should be reflected in some way by the model formula. If the
   model formula fails to describe the non-linear trend, then problems
   will persist.
3. the model proposed is missing an important covariate that might
   help standardise the data in a way that results in linearity


:::

::::

- **independence**: the residuals (and thus observations) must be
  independently drawn from the populations. That is, the correlation
  between all the observations is assumed to be 0 (off-diagonals in
  the covariance matrix). More practically, there should be no
  pattern to the correlations between observations.

  Random sampling and random treatment assignment are experimental
  design elements that are intended to mitigate many types of sampling
  biases that cause dependencies between observations. Nevertheless,
  there are aspects of sampling designs that are either logistically
  difficult to randomise or in some cases not logically possible. For
  example, the residuals from observations sampled closer together in
  space and time will likely be more similar to one another than those
  of observations more spaced apart. Since neither space nor time can
  be randomised, data collected from sampling designs that involve
  sampling over space and/or time need to be assess for spatial and
  temporal dependencies. These concepts will be explored in the
  context of introducing susceptible designs in a later tutorial.

The above is only a very brief overview of the model assumptions that
apply to just one specific model (simple linear gaussian regression).
For the remainder of this section, we will graphically explore the two
motivating example data sets so as gain insights into what
distributional assumptions might be most valid, and thus help guide
modelling choices. Similarly, for subsequent tutorials in this series
(that introduce progressively more complex models), all associated
assumptions will be explored and detailed.

::: {.panel-tabset}

## Example 1 (Gaussian data)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda1a
#| fig-width: 3
#| fig-height: 3
dat |> 
  ggplot(aes(y = y)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- there is no strong evidence of non-normality
- to be convincing evidence of non-normality, each segment of the
  boxplot should get progressively larger

:::::: 
::::: 

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda1b
#| fig-width: 3
#| fig-height: 3
dat |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of values around the trendline seems fairly even (hence
  it there is no evidence of non-homogeneity

:::::: 
::::: 

### Linearity

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda1c
#| fig-width: 3
#| fig-height: 3
dat |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the data seems well represented by the linear trendline.
  Furthermore, the lowess smoother does not appear to have a
  consistent shift trajectory.

:::::: 
::::: 
::::


**Conclusions**

- there are no obvious violations of the linear regression model assumptions
- we can now fit the suggested model
- full confirmation about the model's goodness of fit should be
  reserved until after exploring the additional diagnostics that are
  only available after fitting the model.


## Example 2 (categorical predictor)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda2a
#| fig-width: 3
#| fig-height: 3
dat2 |> 
  ggplot(aes(y = y, x = x)) +
  geom_boxplot()
```
::::::
:::::: {.column width="48%"}
**Conclusions**

- there is no consitent evidence of non-normality across all groups
- even though the control group demonstrates some evidence of
  non-normality
:::::: 
:::::

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda2b
#| fig-width: 3
#| fig-height: 3
dat2 |> 
  ggplot(aes(y = y, x = x)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of noise in each group seems reasonably similar
- more importantly, there does not seem to be a relationship between
  the mean (as approximated by the position of the boxplots along the
  y-axis) and the variance (as approximated by the spread of the
  boxplots).
- that is, the size of the boxplots do not vary with the elevation of
  the boxplots.

:::::: 
::::: 

### Linearity

Linearity is not an issue for categorical predictors since it is
effectively fitting separate lines between pairs of points (and a line
between two points can only ever be linear)....

::::

**Conclusions**

- no evidence of non-normality
- no evidence of non-homogeneity of variance


## Example 3 (Poisson data)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda3a
#| fig-width: 3
#| fig-height: 3
dat3 |> 
  ggplot(aes(y = y)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- there is strong evidence of non-normality
- each segment of the boxplot should get progressively larger

:::::: 
::::: 

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda3b
#| fig-width: 3
#| fig-height: 3
dat3 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of noise does not look random along the line of best fit.
- homogeneity of variance is difficult to assess in the presence of
  distributional issues (such as non-normality in this case) as they
  can result in non-linearity (apparent here)

:::::: 
::::: 

### Linearity

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda3c
#| fig-width: 3
#| fig-height: 3
dat3 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the data do not appear to be linear

:::::: 
::::: 
::::


**Conclusions**

- there are obvious violations of the linear regression model assumptions
- we should consider a different model that does not assume normality


## Example 4 (NB data)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda4a
#| fig-width: 3
#| fig-height: 3
dat4 |> 
  ggplot(aes(y = y)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- there is strong evidence of non-normality
- each segment of the boxplot should get progressively larger

:::::: 
::::: 

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda4b
#| fig-width: 3
#| fig-height: 3
dat4 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of noise does not look random along the line of best fit.
- homogeneity of variance is difficult to assess in the presence of
  distributional issues (such as non-normality in this case) as they
  can result in non-linearity (apparent here)

:::::: 
::::: 

### Linearity

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda4c
#| fig-width: 3
#| fig-height: 3
dat4 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the data do not appear to be linear

:::::: 
::::: 
::::


**Conclusions**

- there are obvious violations of the linear regression model assumptions
- we should consider a different model that does not assume normality

## Example 5 (Binary data)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda5a
#| fig-width: 3
#| fig-height: 3
dat5 |> 
  ggplot(aes(y = y)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- clearly a set of 0's and 1's cant be normally distributied.

:::::: 
::::: 

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda5b
#| fig-width: 3
#| fig-height: 3
dat5 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of noise does not look random (or equal) along the line
  of best fit.

:::::: 
::::: 

### Linearity

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda5c
#| fig-width: 3
#| fig-height: 3
dat5 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the data are clearly not linear

:::::: 
::::: 
::::


**Conclusions**

- there are obvious violations of the linear regression model assumptions
- we should consider a different model that does not assume normality

## Example 6 (Binomial data)

:::: {.panel-tabset}

### Normality

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda6a
#| fig-width: 3
#| fig-height: 3
dat6 |> 
  ggplot(aes(y = y)) +
  geom_boxplot()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- distribution is not normal and is truncated

:::::: 
::::: 

### Homogeneity of variance

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda6b
#| fig-width: 3
#| fig-height: 3
dat6 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- the spread of noise does not look random (or equal) along the line
  of best fit.

:::::: 
::::: 

### Linearity

::::: {.columns}
:::::: {.column width="48%"}

```{r}
#| label: eda6c
#| fig-width: 3
#| fig-height: 3
dat6 |> 
  ggplot(aes(y = y, x = x)) +
  geom_smooth(method = "lm") +
  geom_point()
```

:::::: 

:::::: {.column width="48%"}
**Conclusions**

- although there is no evidence of non-linearity from this small data
  set, it is worth noting that the line of best fit does extend
  outside the logical response range [0.1] within the range of
  observed $x$ values. That is, a simple linear model would predict
  proportions higher than 100% at high values of $x$

:::::: 
::::: 
::::


**Conclusions**

- there are obvious violations of the linear regression model assumptions
- we should consider a different model that does not assume normality

:::

# Fitting models


:::: {.panel-tabset}


### Example 1 (Gaussian data)

::::::: {.panel-tabset}
#### Raw predictor
::::: {.panel-tabset}

##### Define priors

One way to assess the priors is to have the MCMC sampler sample purely
from the prior predictive distribution without conditioning on the
observed data. Doing so provides a glimpse at the range of predictions
possible under the priors. On the one hand, wide ranging predictions
would ensure that the priors are unlikely to influence the actual
predictions once they are conditioned on the data. On the other hand,
if they are too wide, the sampler is being permitted to traverse into
regions of parameter space that are not logically possible in the
context of the actual underlying ecological context. Not only could
this mean that illogical parameter estimates are possible, when the
sampler is traversing regions of parameter space that are not
supported by the actual data, the sampler can become unstable and have
difficulty.

In `brms`, we can inform the sampler to draw from the prior predictive
distribution instead of conditioning on the response, by running the
model with the `sample_prior = 'only'` argument. Unfortunately, this
cannot be applied when there are flat priors (since the posteriors
will necessarily extend to negative and positive infinity). Therefore,
in order to use this useful routine, we need to make sure that we have
defined a proper prior for all parameters.

Earlier we suggested the following priors might be useful:

- $\beta_0$: Normal prior centred at `r round(median(dat$y),2)` with a
  variance of `r round(mad(dat$y),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(dat$y)/mad(dat$x),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```
 
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(mad(dat$y),2)`
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```

It might be use usefull to understand what some of these distributions
look like. For example, we have used both a normal (Gaussian)
distribution and a flatter _t_ distribution for y-intercept and slope
respectively. This was a somewhat arbitrary choice. We could easily
have gone with either normal or _t_ distributions for all of the above
parameters. To visualise prior distributions for the slope based on
both normal and _t_ distributions:

```{r} 
#| label: normal_prior
#| eval: true
#| echo: true
#| fig-width: 4
#| fig-height: 3
standist::visualize("normal(0, 4.09)", "student_t(3, 0, 4.09)", xlim = c(-20, 20))
```

Evidently, the _t_ distribution (with 3 degrees of freedom) is wider
than the normal distribution. The former should be more robust to data
with values that are less concentrated around the mean.

```{r} 
#| label: set_priors_1a
#| eval: true
#| echo: true
priors <- prior(normal(31.21, 15.17), class = "Intercept") +
    prior(student_t(3, 0, 4.09), class = "b") +
    prior(student_t(3, 0, 15.17), class = "sigma")
```


##### Fit the model
$$
\begin{align}
y_i \sim{}& N(\mu_i, \sigma^2)\\
\mu_i =& \beta_0 + \beta_1 x_i\\
\end{align}
$$

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms1a
#| eval: true
#| echo: true
#| cache: true
dat1a.brm <- brm(bf(y ~ x),
                data=dat,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                #backend = 'cmdstan',
                refresh = 0)
```

:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_1a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1a.brm |>
    conditional_effects() |>
    plot(points = TRUE)
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.

:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_1a
#| eval: true
#| echo: true
#| cache: true
dat1a.brm2 <- update(dat1a.brm,
    sample_prior = "yes"
)
```


:::


4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_1a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1a.brm2 |>
    conditional_effects() |>
    plot(points = TRUE)
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_1a
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1a.brm2 |> tidybayes::get_variables()
dat1a.brm2 |> hypothesis("x = 0") |> plot()
dat1a.brm2 |> hypothesis("sigma = 0", class = "") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel2k, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat1a.brm2 |> tidybayes::get_variables()
dat1a.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

#### Centered predictor

::::: {.panel-tabset}

##### Define priors

Unless you explicitly direct `brm` to include a user-defined
intercept, the priors on the default intercept should assume that the
predictor(s) are centered (because `brm` will automatically center all
continuous predictors).

Lets try the following priors:

- $\beta_0$: Normal prior centred at `r round(median(dat$y),2)` with a
  variance of `r round(mad(dat$y),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_1b
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> median() |> round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_1b
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> mad() |> round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centered at
  0 with a variance of `r round(mad(dat$y)/mad(dat$x),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_1b
    #| echo: true
    #| eval: true
    #| cache: false
    mad(dat$y) / mad(dat$x) |> round(2)
    ```
 
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(mad(dat$y),2)`
  - variance: 
    ```{r}
    #| prior_y_var_1b
    #| echo: true
    #| eval: true
    #| cache: false
    ```

```{r} 
#| label: set_priors_1b
#| eval: true
#| echo: true
priors <- prior(normal(31.21, 15.17), class = "Intercept") +
    prior(student_t(3, 0, 4.09), class = "b") +
    prior(student_t(3, 0, 15.17), class = "sigma")
```


##### Fit the model
$$
\begin{align}
y_i \sim{}& N(\mu_i, \sigma^2)\\
\mu_i =& \beta_0 + \beta_1 (x_i - \bar{x})\\
\end{align}
$$

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms1b
#| eval: true
#| echo: true
#| cache: true
dat1b.brm <- brm(bf(y ~ scale(x, scale = FALSE)),
                data=dat,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```

:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_1b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1b.brm |>
    conditional_effects() |>
    plot(points = TRUE)
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.

:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_1b
#| eval: true
#| echo: true
#| cache: true
dat1b.brm2 <- update(dat1b.brm,
    sample_prior = "yes"
)
```


:::


4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_1b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1b.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_1b
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1b.brm2 |> tidybayes::get_variables() 
dat1b.brm2 |> hypothesis("scalexscaleEQFALSE = 0") |> plot()
dat1b.brm2 |> hypothesis("sigma = 0", class = "") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel1b, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat1b.brm2 |> tidybayes::get_variables() 
dat1b.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

#### Standardised predictor

::::: {.panel-tabset}

##### Define priors

When the predictor is standardised, it simplifies prior definition
because we no longer need to consider the scale of the predictor.

Lets try the following priors:

- $\beta_0$: Normal prior centred at `r round(median(dat$y),2)` with a
  variance of `r round(mad(dat$y),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_1c
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> median() |> round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_1c
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> mad() |> round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centered at
  0 with a variance of `r round(mad(dat$y),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_1c
    #| echo: true
    #| eval: true
    #| cache: false
    mad(dat$y) |> round(2)
    ```
 
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(mad(dat$y),2)`
  - variance: 
    ```{r}
    #| prior_y_var_1c
    #| echo: true
    #| eval: true
    #| cache: false
    ```

```{r} 
#| label: set_priors_1c
#| eval: true
#| echo: true
priors <- prior(normal(31.21, 15.17), class = "Intercept") +
    prior(student_t(3, 0, 15.17), class = "b") +
    prior(student_t(3, 0, 15.17), class = "sigma")
```


##### Fit the model

$$
\begin{align}
y_i \sim{}& N(\mu_i, \sigma^2)\\
\mu_i =& \beta_0 + \beta_1 (x_i - \bar{x})/\sigma_{x}\\
\end{align}
$$

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms1c
#| eval: true
#| echo: true
#| cache: true
dat1c.brm <- brm(bf(y ~ scale(x)),
                data=dat,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```

:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_1c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1c.brm |>
    conditional_effects() |>
    plot(points = TRUE)
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.

:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_1c
#| eval: true
#| echo: true
#| cache: true
dat1c.brm2 <- update(dat1c.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_1c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1c.brm2 |>
    conditional_effects() |>
    plot(points = TRUE)
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_1c
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat1c.brm2 |> tidybayes::get_variables()
dat1c.brm2 |> hypothesis("scalex = 0") |> plot()
dat1c.brm2 |> hypothesis("sigma = 0", class = "") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel1c, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat1c.brm2 |> tidybayes::get_variables()
dat1c.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

:::::::

### Example 2 (categorical predictor)

::::::: {.panel-tabset}
#### Treatment contrasts
::::: {.panel-tabset}

##### Define priors

Lets try the following priors:

```{r}
#| label: prior_y_means_2a
#| echo: FALSE
#| cache: FALSE
dat2_prior <- dat2 |>
  group_by(x) |>
  summarise(across(y, list(med = median, sd = sd, mad = mad))) |>
  as.data.frame()

dat2_prior_b <-
  dat2 |>
  group_by(x) |>
  summarise(across(y, median)) |>
  pull(y) |>
  diff() |>
  abs() |>
  max()
dat2_prior_sig <-
  dat2 |>
  group_by(x) |>
  summarise(across(y, mad)) |>
  pull(y) |>
  max()
```

- $\beta_0$: Normal prior centred at `r round(dat2_prior[1,2],2)` with a
  variance of `r round(dat2_prior[1,4],2)`
  - mean: 
    ```{r}
    #| prior_y_mean_2a
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
      group_by(x) |>
        summarise(across(y, list(med = median, sd = sd, mad = mad)))
    ```
  - variance: 
    ```{r}
    #| prior_y_var_2a
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
      group_by(x) |>
        summarise(across(y, list(med = median, sd = sd, mad = mad)))
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centered at
  0 with a variance of `r round(dat2_prior_b,2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_2a
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
      group_by(x) |>
      summarise(across(y, median)) |>
      pull(y) |>
      diff() |>
      abs() |>
      max()
    ```
 
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(dat2_prior_sig, 2)`
  - variance: 
    ```{r}
    #| prior_y_var2_2a
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
    group_by(x) |>
    summarise(across(y, mad)) |>
    pull(y) |>
    max()
    ```

```{r} 
#| label: set_priors_2a
#| eval: true
#| echo: true
priors <- prior(normal(19.68, 1.87), class = "Intercept") +
    prior(student_t(3, 0, 6.35), class = "b") +
    prior(student_t(3, 0, 4.7), class = "sigma")
```

##### Fit the model

$$
\begin{align}
y_i \sim{}& N(\mu_i, \sigma^2)\\
\mu_i =& \beta_0 + \sum{\beta_j x_{ij}}\\
\end{align}
$$

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms2a
#| eval: true
#| echo: true
#| cache: true
dat2a.brm <- brm(bf(y ~ x),
                data=dat2,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0) 
```

:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_2a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2a.brm |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.

:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_2a
#| eval: true
#| echo: true
#| cache: true
dat2a.brm2 <- update(dat2a.brm,
    sample_prior = "yes"
) 
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_2a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2a.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_2a
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2a.brm2 |> tidybayes::get_variables()
dat2a.brm2 |> hypothesis("xmedium = 0") |> plot()
dat2a.brm2 |> hypothesis("sigma = 0", class = "") |> plot() 
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel2a, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat2a.brm2 |> tidybayes::get_variables()
dat2a.brm2 |> SUYR_prior_and_posterior() 
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

#### Means parameterisation

::::: {.panel-tabset}

##### Define priors

Lets try the following priors:

```{r}
#| label: prior_y_means_2b
#| echo: FALSE
#| cache: FALSE
dat2_prior <-
  dat2 |>
  summarise(across(y, list(med = median, sd = sd, mad = mad))) |>
  as.data.frame()

dat2_prior_sig <-
  dat2 |>
  group_by(x) |>
  summarise(across(y, mad)) |>
  pull(y) |>
  max()
```

- $\beta$: _t_ distribution (3 degrees of freedom) prior centered at
  `r round(dat2_prior[1,1], 2)` with a variance of `r round(dat2_prior[1,3],2)`
  - mean: since each groups mean is being estimated separately, they
    could either all have different priors, or more commonly, the same
    priors.
  - variance: 
    ```{r}
    #| prior_y_var1_2b
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
        summarise(across(y, list(med = median, sd = sd, mad = mad)))
    ```
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(dat2_prior_sig, 2)`
  - variance: 
    ```{r}
    #| prior_y_var2_2b
    #| echo: true
    #| eval: true
    #| cache: false
    dat2 |>
    group_by(x) |>
    summarise(across(y, mad)) |>
    pull(y) |>
    max()
    ```

```{r} 
#| label: set_priors_2b
#| eval: true
#| echo: true
priors <- prior(student_t(3, 16.18, 6.56), class = "b") +
    prior(student_t(3, 0, 4.7), class = "sigma")
```

##### Fit the model

$$
\begin{align}
y_i \sim{}& N(\mu_i, \sigma^2)\\
\mu_i =& \beta_0 + \sum{\beta_j x_{ij}}\\
\end{align}
$$

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms2b
#| eval: true
#| echo: true
#| cache: true
dat2b.brm <- brm(bf(y ~ -1 + x),
                data=dat2,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0) 
```

:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_2b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2b.brm |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.

:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_2b
#| eval: true
#| echo: true
#| cache: true
dat2b.brm2 <- update(dat2b.brm,
    sample_prior = "yes"
) 
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_2b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2b.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_2b
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat2b.brm2 |> tidybayes::get_variables()
dat2b.brm2 |> hypothesis("xcontrol = 0") |> plot()
dat2b.brm2 |> hypothesis("xmedium = 0") |> plot()
dat2b.brm2 |> hypothesis("sigma = 0", class = "") |> plot() 
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel2b, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat2b.brm2 |> tidybayes::get_variables()
#dat2b.brm2 |> SUYR_prior_and_posterior() 
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::
:::::::

### Example 3 (Poisson data)

::::::: {.panel-tabset}
#### Raw predictor

$$
\begin{align}
y_i \sim{}& Pois(\lambda_i)\\
log(\lambda_i) =& \beta_0 + \beta_1 x_i\\
\end{align}
$$

::::: {.panel-tabset}

##### Define priors

When considering priors, it is important to remember that they apply
to parameters on the link scale. For Poisson models, the link scale is
log. So the following priors might be appropriate:

- $\beta_0$: Normal prior centred at `r round(median(log(dat3$y)),2)` with a
  variance of `r round(mad(log(dat3$y)),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_3a
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        median() |>
        round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_3a
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        mad() |>
        round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(log(dat3$y))/mad(log(dat3$x)),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_3a
    #| echo: true
    #| eval: true
    #| cache: false
    dat3 |>
        mutate(across(c(y, x), log)) |>
        summarise(across(c(y, x), mad)) |>
        mutate(round(y / x, 2))
    ```
 
```{r} 
#| label: set_priors_3a
#| eval: true
#| echo: true
priors <- prior(normal(2.00, 1.33), class = "Intercept") +
    prior(student_t(3, 0, 2.00), class = "b")
```

##### Fit the model

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms3a
#| eval: true
#| echo: true
#| cache: true
dat3a.form <- bf(y ~ x, family = poisson(link = "log"))
dat3a.brm <- brm(dat3a.form,
                data=dat3,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```
:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_3a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3a.brm |>
  conditional_effects() |>
  plot(points = TRUE) |> _[[1]] +
  scale_y_log10()
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.


:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_3a
#| eval: true
#| echo: true
#| cache: true
dat3a.brm2 <- update(dat3a.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_3a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3a.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_3a
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3a.brm2 |> tidybayes::get_variables() 
dat3a.brm2 |> hypothesis("x = 0") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel3a, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat3a.brm2 |> tidybayes::get_variables() 
dat3a.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::


#### Centered predictor

$$
\begin{align}
y_i \sim{}& Pois(\mu_i)\\
log(\mu_i) =& \beta_0 + (\beta_1 x_i - \bar{x})\\
\end{align}
$$

::::: {.panel-tabset}

##### Define priors

When considering priors, it is important to remember that they apply
to parameters on the link scale. For Poisson models, the link scale is
log. So the following priors might be appropriate:

- $\beta_0$: Normal prior centred at `r round(median(log(dat3$y)),2)` with a
  variance of `r round(mad(log(dat3$y)),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_3b
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        median() |>
        round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_3b
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        mad() |>
        round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(log(dat3$y))/mad(log(dat3$x)),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_3b
    #| echo: true
    #| eval: true
    #| cache: false
    dat3 |>
        mutate(across(c(y, x), log)) |>
        summarise(across(c(y, x), mad)) |>
        mutate(round(y / x, 2))
    ```
 
```{r} 
#| label: set_priors_3b
#| eval: true
#| echo: true
priors <- prior(normal(2.00, 1.33), class = "Intercept") +
    prior(student_t(3, 0, 2.00), class = "b")
```

##### Fit the model

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms3b
#| eval: true
#| echo: true
#| cache: true
dat3b.form <- bf(y ~ scale(x, scale = FALSE), family = poisson(link = "log"))
dat3b.brm <- brm(dat3b.form,
                data=dat3,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```
:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_3b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3b.brm |>
  conditional_effects() |>
  plot(points = TRUE) |> _[[1]] +
  scale_y_log10()
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.


:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_3b
#| eval: true
#| echo: true
#| cache: true
dat3b.brm2 <- update(dat3b.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_3b
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3b.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_3b
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3b.brm2 |> tidybayes::get_variables() 
dat3b.brm2 |> hypothesis("scalexscaleEQFALSE = 0") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel3b, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat3b.brm2 |> tidybayes::get_variables() 
dat3b.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::


#### Standardised predictor

$$
\begin{align}
y_i \sim{}& Pois(\lambda_i)\\
log(\lambda_i) =& \beta_0 + (\beta_1 x_i - \bar{x})/\sigma_x\\
\end{align}
$$

::::: {.panel-tabset}

##### Define priors

When considering priors, it is important to remember that they apply
to parameters on the link scale. For Poisson models, the link scale is
log. So the following priors might be appropriate:

- $\beta_0$: Normal prior centred at `r round(median(log(dat3$y)),2)` with a
  variance of `r round(mad(log(dat3$y)),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_3c
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        median() |>
        round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_3c
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |>
        log() |> 
        mad() |>
        round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(log(dat3$y)),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_3c
    #| echo: true
    #| eval: true
    #| cache: false
    dat3$y |> log() |> mad() |> round(2)
    ```
 
```{r} 
#| label: set_priors_3c
#| eval: true
#| echo: true
priors <- prior(normal(2.00, 1.33), class = "Intercept") +
    prior(student_t(3, 0, 1.33), class = "b")
```

##### Fit the model

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms3c
#| eval: true
#| echo: true
#| cache: true
dat3c.form <- bf(y ~ scale(x), family = poisson(link = "log"))
dat3c.brm <- brm(dat3c.form,
                data=dat3,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```
:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_3c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3c.brm |>
  conditional_effects() |>
  plot(points = TRUE) |> _[[1]] +
  scale_y_log10()
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.


:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_3c
#| eval: true
#| echo: true
#| cache: true
dat3c.brm2 <- update(dat3c.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_3c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3c.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_3c
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat3c.brm2 |> tidybayes::get_variables() 
dat3c.brm2 |> hypothesis("scalex = 0") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel3c, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat3c.brm2 |> tidybayes::get_variables() 
dat3c.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

:::::::

### Example 5 (Binary data)

::::::: {.panel-tabset}
#### Raw predictor

$$
\begin{align}
y_i \sim{}& Bin(\pi_i, 1)\\
log(\frac{\pi_i}{1-\pi_i}) =& \beta_0 + \beta_1 x_i\\
\end{align}
$$

::::: {.panel-tabset}

##### Define priors

When considering priors, it is important to remember that they apply
to parameters on the link scale. For Binary models, the link scale is
logit. Binomial data is notoriously difficult to define priors for.  Nevertheless the following considerations are useful:

- the observed response values are only ever either 0 or 1
- a linear model is exploring whether the probability of a 1 changes from high to low or low to high according to the linear predictor
- the switch in probability is likely to be somewhere near the middle of the $x$ range
- with a centered predictor, the mean response is expected to be approximately 0.5
- on a logit (log odds) scale, this corresponds to a value of 0.
- on a logit (log odds) scale, values of -3 and 3 are considered very wide
- on a logit scale, values between -1 and 1 are reasonable.

So the following priors might be appropriate:

- $\beta_0$: Normal prior centred at 0 with a
variance of 1
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of 1

```{r} 
#| label: set_priors_5a
#| eval: true
#| echo: true
priors <- prior(normal(0, 1), class = "Intercept") +
    prior(student_t(3, 0, 1), class = "b")
```

##### Fit the model

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms5a
#| eval: true
#| echo: true
#| cache: true
dat5a.form <- bf(y ~ x, family = binomial(link = "logit"))
dat5a.brm <- brm(dat5a.form,
                data=dat5,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```
:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

For Binary data, it is often more useful to explore the predictions on
the link scale. Ribbons that extend much beyond -3 and 3 would
definitely be considered wide enough.

```{r}
#| label: brms_conditional_effects_5a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5a.brm |>
  conditional_effects(method = "posterior_linpred") 

dat5a.brm |>
  conditional_effects() |>
  plot(points = TRUE) 
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.


:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_5a
#| eval: true
#| echo: true
#| cache: true
dat5a.brm2 <- update(dat5a.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_5a
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5a.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_5a
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5a.brm2 |> tidybayes::get_variables() 
dat5a.brm2 |> hypothesis("x = 0") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel5a, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat5a.brm2 |> tidybayes::get_variables() 
dat5a.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

#### Standardised predictor

$$
\begin{align}
y_i \sim{}& Bin(\pi_i, 1)\\
log(\frac{\pi_i}{1 -\pi_i}) =& \beta_0 + (\beta_1 x_i - \bar{x})/\sigma_x\\
\end{align}
$$

::::: {.panel-tabset}

##### Define priors

- $\beta_0$: Normal prior centred at 0 with a
variance of 1
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of 1

```{r} 
#| label: set_priors_5c
#| eval: true
#| echo: true
priors <- prior(normal(0, 1), class = "Intercept") +
    prior(student_t(3, 0, 1), class = "b")
```

##### Fit the model

1. start by fitting the model and sampling from the priors only

::: {.indented}

```{r}
#| label: brms5c
#| eval: true
#| echo: true
#| cache: true
dat5c.form <- bf(y ~ scale(x), family = binomial(link = "logit"))
dat5c.brm <- brm(dat5c.form,
                data=dat5,
                prior=priors,
                sample_prior = 'only', 
                iter = 5000,
                warmup = 1000,
                chains = 3, cores = 3,
                thin = 5,
                backend = 'cmdstan',
                refresh = 0)
```
:::

2. explore the range of posterior predictions resulting from the
   priors alone

::: {.indented}

```{r}
#| label: brms_conditional_effects_5c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5c.brm |>
  conditional_effects(method = "posterior_linpred") 
```

**Conclusions:** 

- the grey ribbon above represents the credible range of the posterior
  predictions
- this ribbon is substantially wider than the observed data (black
  points), suggesting that the priors are not overly narrow
- the range of posterior predictions are not widely unreasonable
  (although we could argue that negative predictions might be
  illogical) since the maximums are not multiple orders of magnitude
  above the observed data.


:::

3. now refit the model such that it samples from both the priors and
   likelihood (that is allow the data to have an impact on the
   estimates)

::: {.indented}

```{r}
#| label: brms_fit2_5c
#| eval: true
#| echo: true
#| cache: true
dat5c.brm2 <- update(dat5c.brm,
    sample_prior = "yes"
)
```

:::

4. re-explore the range of posterior predictions resulting from the
   fitted model

::: {.indented}

```{r}
#| label: brms_conditional_effects2_5c
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5c.brm2 |>
    conditional_effects() |>
    plot(points = TRUE) 
```

**Conclusions:** 

- the range of the posteriors are now substantially reduced (now that
  the model includes both data and priors)
- this suggests that the patterns are being driven predominantly by the data

:::

5. compare the priors and posteriors to further confirm that the
   priors are not overly influential
   
::: {.indented}

When we have indicated that the posterior should be informed by both
the prior and the posterior, both prior (governed by priors alone) and
posterior (governed by both priors and data/likelihood) draws are
returned.  These can be compared by exploring the probabilities
associated with specific hypotheses - the most obvious of which is
that of no effect (that the parameter = 0).

When doing so, ideally the posterior should be distinct from the
prior. By distinct, I mean that the posteriors should be clearly
distinguishable from the priors. Ideally, the priors should be less
variable than the priors.  If this is not the case, it might suggest
that the posteriors are being too strongly driven by the priors.

```{r}
#| labels: brms_prior_posterior_5c
#| echo: true
#| eval: true
#| cache: true
#| fig-width: 4
#| fig-height: 3
dat5c.brm2 |> tidybayes::get_variables() 
dat5c.brm2 |> hypothesis("scalex = 0") |> plot()
```

Unfortunately, it is not possible to do this comparison sensibly for
the intercept.  The reason for this is that the prior for intercept
was applied to an intercept that is associated with centred
continuous predictors (predictors are centred behind the scenes).
Since we did not centre the predictor, the intercept returned is as if
uncentred.  Hence, the prior and posterior are on different scales.
 
```{r fitModel5c, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 8, fig.height = 4}
dat5c.brm2 |> tidybayes::get_variables() 
dat5c.brm2 |> SUYR_prior_and_posterior()
```

**Conclusions:**

- each of the priors are substantially wider than the posteriors
- the posteriors are definitely distinct from their respective priors
  and thus we can conclude that the priors and not driving the
  posteriors (e.g. they are not influencing the outcomes)
- the priors are simply regularising the parameters such that they are
  only sampled from plausible regions

:::

:::::

:::::::

::::

# References

