---
title: Bayesian linear models
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    toc-float: true
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    html-math-method: katex
execute:
  keep-md: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: resources/references.bib
output_dir: "docs"
---


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)   #for data wrangling and plotting
library(DHARMa)      #for simulated residuals
library(performance) #for model diagnostics
library(see)         #for model diagnostics
library(brms)        #for Bayesian models
library(tidybayes)   #for exploring Bayesian PKPDmodels
library(rstan)       #for diagnostics plots
library(bayesplot)   #for diagnostic plots
library(patchwork)   #for arranging multiple plots
library(HDInterval)  #for HPD intervals
library(bayestestR)  #for ROPE
library(emmeans)     #for estimated marginal means
source("helperfunctions.R")
```

Many biologists and ecologists get a little twitchy and nervous around
mathematical and statistical formulae and nomenclature. Whilst it is
possible to perform basic statistics without too much regard for the
actual equation (model) being employed, as the complexity of the
analysis increases, the need to understand the underlying model
becomes increasingly important. Moreover, model specification in BUGS
(the language used to program Bayesian modelling) aligns very closely
to the underlying formulae. Hence a good understanding of the
underlying model is vital to be able to create a sensible Bayesian
model. Consequently, I will always present the linear model formulae
along with the analysis. If you start to feel some form of disorder
starting to develop, you might like to run through the Tutorials and
Workshops twice (the first time ignoring the formulae). 

# A philosophical note

To introduce the philosophical and mathematical differences between
classical (frequentist) and Bayesian statistics, @Wade-2000 presented
a provocative yet compelling trend analysis of two hypothetical
populations. The temporal trend of one of the populations shows very
little variability from a very subtle linear decline. By contrast, the
second population appears to decline more dramatically, yet has
substantially more variability.

@Wade-2000 neatly illustrates the contrasting conclusions
(particularly with respect to interpreting probability) that would be
drawn by the frequentist and Bayesian approaches and in so doing
highlights how and why the Bayesian approach provides outcomes that
are more aligned with management requirements.

This tutorial will start by replicating the demonstration of @Wade-2000.

```{r}
#| label: Pops
#| eval: true
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| tidy: true
set.seed(1)
x <- 1:10
b <- -0.03
a <- 200
sigma <- 0.11
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -0.1
a <- 200
sigma <- 0.38
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popA <- data.frame(y,x)

set.seed(1)
x <- 1:10
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x)
#summary(lm(y~x))
popB <- data.frame(y,x)
#plot(y~x)

set.seed(1)
x <- seq(1,10,l=100)
b <- -10
a <- 200
sigma <- 40
mm <- model.matrix(~x)
y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
#plot(y~x, ylim=c(0,150))
#summary(lm(y~x))
popC <- data.frame(y,x)
#plot(y~x)

#par(mar=c(3,3,0,0),mfrow=c(1,3))
#plot(y~x,popA, ylim=c(0,250))
#abline(lm(y~x,popA))
#plot(y~x,popB, ylim=c(0,250))
#abline(lm(y~x,popB))
#plot(y~x,popC, ylim=c(0,250))
#abline(lm(y~x,popC))

##THIS IS BETTER
for (n in 1:1000) {
  set.seed(n)
  x <- 1:10
  b <- -0.1
  a <- 200
  sigma <- 0.38
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popA <- data.frame(y,x))
  
  set.seed(n)
  x <- 1:10
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  (popB <- data.frame(y,x))
  
  set.seed(n)
  x <- seq(1,10,l=100)
  b <- -10
  a <- 200
  sigma <- 40
  mm <- model.matrix(~x)
  y <- (mm %*% c(a,b))+rnorm(length(x),0,sigma)
  popC <- data.frame(y,x)
  
  aa<-popA.lm <- lm(y~x,popA)
  bb<-popB.lm <- lm(y~x,popB)
  cc<-popC.lm <- lm(y~x,popC)
  AA<-summary(popA.lm)$coef
  BB<-summary(popB.lm)$coef
  CC<-summary(popC.lm)$coef
  
  if(AA[2,4]<0.05 & BB[2,4]>0.05 & abs(BB[2,1]-CC[2,1])<1) break
}
```
::: {.columns}

:::: {.column width="32%"}
```{r}
#| label: PopA
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popA, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popA))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popA.lm <- lm(y~x,popA)
popA.sum<-summary(popA.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popA)`<br>
Slope: `r round(coef(popA.lm)[2],4)`<br>
_t_: `r round(popA.sum[2,3],4)`<br>
_p-value_: `r round(popA.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopB
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
#| tidy: true
par(mar=c(3,3,0,0))
plot(y~x,popB, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popB))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popB.lm <- lm(y~x,popB)
popB.sum<-summary(popB.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popB)`<br>
Slope: `r round(coef(popB.lm)[2],4)`<br>
_t_: `r round(popB.sum[2,3],4)`<br>
_p-value_: `r round(popB.sum[2,4],4)`<br>
</p>

::::

:::: {.column width="32%"}
```{r}
#| label: PopC
#| eval: true
#| echo: false
#| fig-width: 3
#| fig-height: 3
par(mar=c(3,3,0,0))
plot(y~x,popC, ylim=c(0,250), axes=F, ann=F, pch=16)
abline(lm(y~x,popC))
axis(1)
axis(2, las=1)
mtext("x",1,line=2, cex=1.5)
box(bty="l")
popC.lm <- lm(y~x,popC)
popC.sum<-summary(popC.lm)$coef
```

<p style="margin-left:30pt;">
n: `r  nrow(popC)`<br>
Slope: `r round(coef(popC.lm)[2],4)`<br>
_t_: `r round(popC.sum[2,3],4)`<br>
_p-value_: `r round(popC.sum[2,4],4)`<br>
</p>

::::

:::


From a traditional frequentist perspective, we would conclude that
there is a 'significant' relationship in Population A and C
($p<0.05$), yet not in Population B ($p>0.05$). Note, Population B and
C were both generated from the same random distribution, it is just
that Population C has a substantially higher number of observations.

The above illustrates a couple of things

- statistical significance does not necessarily translate into
  biological importance. The percentage of decline for Population A is
  `r p1 <- predict(popA.lm, data.frame( x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)` where as the percentage of decline
  for Population B is `r p1 <- predict(popB.lm, data.frame(x=c(1,10)));
  round(100*(p1[1]-p1[2])/p1[1],2)`. That is Population B is declining
  at nearly 10 times the rate of Population A. That sounds rather
  important, yet on the basis of the hypothesis test, we would dismiss
  the decline in Population B.

- that a _p_-value is just the probability of detecting an effect or
  relationship - what is the probability that the sample size is large
  enough to pick up a difference.

Let us now look at it from a Bayesian perspective. I will just provide
the posterior distributions (densities scaled to 0-1 so that they can
be plotted together) for the slope for each population.

```{r}
#| label: Pop_densities
#| eval: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| warning: false
#| message: false
#| results: hide
#| cache: true
## PopA
popA.brm <- brm(y ~ x, data = popA)
slopeA <- popA.brm |>
    as_draws_df() |>
    pull("b_x")
## PopB
popB.brm <- brm(y ~ x, data = popB)
slopeB <- popB.brm |>
    as_draws_df() |>
    pull("b_x")
## PopC
popC.brm <- brm(y ~ x, data = popC)
slopeC <- popC.brm |>
    as_draws_df() |>
    pull("b_x")

ggplot() +
    geom_density(data = NULL, aes(x = slopeA, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeB, y = ..scaled..)) +
    geom_density(data = NULL, aes(x = slopeC, y = ..scaled..)) +
    scale_y_continuous("Relative probability", limits = c(0, 1), expand = c(0, 0)) +
    scale_x_continuous("Regression slope") +
  theme_classic() +
  theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
```

Focusing on Populations A and B, we would conclude:

- the mean (plus or minus CI) slopes for Population A and B are 
  `r round(mean(slopeA),2)` (`r paste(round(quantile(slopeA,
  p=c(0.025,0.975)),2),collapse=",")`) and `r round(mean(slopeB),2)`
  (`r paste(round(quantile(slopeB,
  p=c(0.025,0.975)),2),collapse=",")`) respectively.

- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.
- the Bayesian approach allows us to query the posterior distribution
  is many other ways in order to ask sensible biological questions.
  For example, we might consider that a rate of change of 5% or
  greater represents an important biological impact. For Population A
  and B, the probability that the rate is 5% or greater is 
  `r round(mean(slopeA< -5),2)` and `r round(mean(slopeB < -5),2)`
  respectively.

# Review of (generalised) linear models

I would highly recommend reviewing the information in [the tutorial on
generalised linear models](30_glm.html), particularly the sections
describe linear models, assumption checking and generalised linear
models (GLM). Whilst there are philosophical differences between
frequentist and Bayesian statistics that have implications for how
models are fit and interpreted, model choice and assumption checking
principles are common between the two approaches. Hence, many of these
topics will be assumed, and not fully described in the current
tutorial.

Recall from [the tutorial on generalised linear models](30_glm.html)
that simple linear regression is a linear modelling process that
models a single response variable against one or more predictors with
a linear combination of coefficients and (in the case of a Gaussian
model) can be expressed as:

::: {.columns}

:::: {.column width="48%"}
$$y_i = \beta_0+ \beta_1 x_i+\epsilon_i \hspace{1cm}\epsilon\sim{}N(0,\sigma^2)$$

where:

- $y_i$ is the response value for each of the $i$ observations
- $\beta_0$ is the y-intercept (value of $y$ when $x=0$)
- $\beta_1$ is the slope (rate of chance in $y$ per unit chance in
  $x$)
- $x_i$ is the predictor value for each of the $i$ observations
- $\epsilon_i$ is the residual value of each of the $i$ observations.
  A residual is the difference between the observed value and the
  value expected by the model.

- $\epsilon\sim{}N(0,\sigma^2)$ indicates that the residuals are
  normally distributed with a constant amount of variance

::::

:::: {.column width="48%"}

```{r}
#| label: residuals,
#| eval: true
#| echo: false
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
x=1:10
y=2+5*x
y1=2+5*x + rnorm(10,0,4) 
lm1 <- lm(y1 ~ x)
cc <- coef(lm1)
r <- resid(lm1)
#print(r)
ff <- fitted(lm1)
#print(ff)
g2=ggplot(data=NULL, aes(y=y1, x=x)) + 
    geom_segment(data = NULL, aes(y = ff, x = x, yend = y1, xend = x), colour = "red") +
    geom_point() +
    geom_abline(slope = cc[2], intercept = cc[1], colour = "blue") +
    #annotate(geom='text', y=40, x=2, label=expression(y[i]==2+5*x[i]+epsilon[i]), parse=TRUE, hjust=0) +
    annotate(geom='text', y=30, x=2.5, label=expression(residual~(epsilon[i])), parse=TRUE, hjust=1) +
    annotate(geom='text', y=15, x=5, label=expression(slope~(beta[1])), parse=TRUE, hjust=0) +
    annotate(geom='text', y=-5, x=1, label=expression(y-intercept~(beta[0])), parse=TRUE, hjust=0) +
    geom_curve(aes(x = 2, y = 27, xend = 3.8, yend = 23), 
             curvature = 0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 4.7, y = 15, xend = 3.7, yend = 18), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    geom_curve(aes(x = 0.7, y = -5, xend = -0.5, yend = -0.5), 
             curvature = -0.3, # Adjust curvature as needed
             arrow = arrow(length = unit(0.05, "inches"))) +
    scale_y_continuous('y', limits = c(-10, 50)) +
    xlim(-2, 10) +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
g2
```
::::

:::

The above can be re-expressed and generalised as:
$$
y_i\sim{}Dist(\mu_i, ...)\\
g(\mu_i) = \beta_0+ \beta_1 x_i
$$

where:

- $Dist$ represents a distribution from the exponential family (such
  as Gaussian, Poisson, Binomial, etc)
- $...$ represents additional parameters relevant to the nominated
  distribution (such as $\sigma^2$: Gaussian, $n$: Binomial and
  $\phi$: Negative Binomial, etc)
- $g()$ represents the link function (e.g. log: Poisson, logit:
  Binomial, etc)

The reliability of any model depends on the degree to which the data
adheres to the model assumptions. Hence, as with frequentist models,
exploratory data analysis (EDA) is a vital component of Bayesian
modelling and since the model structures are similar between
frequentist and Bayesian approaches, so too is EDA.

# Bayesian (generalised) linear models

For the purpose of introduction, we will start by exploring a Gaussian
model with a very simple fabricated data set representing the
relationship between a response ($y$) and a continuous predictor ($x =
[1,2,3,4,5,6,7,8,9,10]$. The fabricated data set will comprise 10
observations each drawn from normal distributions with a set standard
deviation of 4. The means of the 10 populations will be determined by
the following equation: 

$$ \mu_i = 2 + 5\times x_i $$

Let generate these data.

```{r}
#| label: sim1
#| eval: true
#| echo: true
#| fig-height: 3
#| fig-width: 3
#| warning: false
set.seed(234)
dat <- data.frame(x = 1:10) |>
    mutate(y = round(rnorm(n = 10, mean = 2 + (5 * x), sd = 4), digits = 2))
dat
```



The model will will be fitting will be:

$$
\begin{align}
y_i&\sim{}N(\mu_i, \sigma^2)\\
\mu_i &= \beta_0+ \beta_1 x_i
\end{align}
$$

The parameters that we are going to attempt to estimate are the
y-intercept ($\beta_0$), the slope ($\beta_1$) and the underlying
variance ($\sigma^2$). Recall (from tutorials on [statistical
philosophies](21_statistical_philosophies.html) and
[estimation](22_estimation.html))) that Bayesian models calculate
**posterior** predictions ($P(H|D)$) from likelihood ($P(D|H)$) and
**prior** expectations ($P(H)$). Therefore, in preparation for fitting
a Bayesian model, we must consider what our prior expectations are for
_all_ parameters.

The individual responses ($y_i$, observed yields) are each expected to
have been **independently** drawn from normal (**Gaussian**)
distributions ($\mathcal{N}$). These distributions represent all the
possible values of $y$ we could have obtained at the specific ($i^{th}$)
level of $x$. Hence the $i^{th}$ $y$ observation is expected
to have been drawn from a normal distribution with a mean of $\mu_i$.

Although each distribution is expected to come from populations that
differ in their means, we assume that all of these distributions have
the **same variance** ($\sigma^2$).

## Priors

We need to supply priors for each of the parameters to be estimated
($\beta_0$, $\beta_1$ and $\sigma$). Whilst we want these priors to be
sufficiently vague as to not influence the outcomes of the analysis
(and thus be equivalent to the frequentist analysis), we do not want
the priors to be so vague (wide) that they permit the MCMC sampler to
drift off into parameter space that is both illogical as well as
numerically awkward.

Proffering sensible priors is one of the most difficult aspects of
performing Bayesian analyses. For instances where there are some
previous knowledge available and a desire to incorporate those data,
the difficulty is in how to ensure that the information is
incorporated correctly. However, for instances where there are no
previous relevant information and so a desire to have the posteriors
driven entirely by the new data, the difficulty is in how to define
priors that are both vague enough (not bias results in their
direction) and yet not so vague as to allow the MCMC sampler to drift
off into unsupported regions (and thus get stuck and yield spurious
estimates).

For early implementations of MCMC sampling routines (such as
Metropolis Hasting and Gibbs), it was fairly common to see very vague
priors being defined.  For example, the priors on effects, were
typically normal priors with mean of 0 and variance of `1e+06`
(1,00,000).  These are very vague priors.  Yet for some samplers
(e.g. NUTS), such vague priors can encourage poor behaviour of the
sampler - particularly if the posterior is complex. It is now
generally advised that priors should (where possible) be somewhat
**weakly informative** and to some extent, represent the bounds of
what are feasible and sensible estimates.

The degree to which priors __influence__ an outcome (whether by having
a pulling effect on the estimates or by encouraging the sampler to
drift off into unsupported regions of the posterior) is dependent on:

- the relative sparsity of the data - the larger the data, the less
  weight the priors have and thus less influence they exert.
- the complexity of the model (and thus posterior) - the more
  parameters, the more sensitive the sampler is to the priors.

The sampled posterior is the product of both the likelihood and the
prior - all of which are multidimensional.  For most applications, it
would be vertically impossible to define a sensible multidimensional
prior.  Hence, our only option is to define priors on individual
parameters (e.g. the intercept, slope(s), variance etc) and to hope
that if they are individually sensible, they will remain collectively
sensible.

So having (hopefully) impressed upon the notion that priors are an
important consideration, I will now attempt to synthesise some of the
approaches that can be employed to arrive at weakly informative priors
that have been gleaned from various sources.  Largely, this advice has
come from the following resources:

- https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations
- http://svmiller.com/blog/2021/02/thinking-about-your-priors-bayesian-analysis/

I will outline some of the current main recommendations before
summarising some approaches in a table.

- weakly informative priors should contain enough information so as to
  regularise (discourage unreasonable parameter estimates whilst
  allowing all reasonable estimates).
- for effects parameters on scaled (standardised) data, an argument
  could be made for a normal distribution with a standard deviation of
  1 (e.g. `normal(0,1)`), although some prefer a _t_ distribution with
  3 degrees of freedom and standard deviation of 1 (e.g.
  `student_t(3,0,1)`) - apparently a flatter _t_ is a more robust
  prior than a normal as an uninformative prior...
- for un-scaled data, the above priors can be scaled by using the
  standard deviation of the data as the prior standard deviation
  (e.g. `student_t(3,0,sd(y))`, or `sudent_t(3,0,sd(y)/sd(x))`)
- for priors of hierachical standard deviations, priors should
  encourage shrinkage towards 0 (particularly if the number of groups
  is small, since otherwise, the sampler will tend to be more
  responsive to "noise").

In this tutorial series, we will perform Bayesian analysis in the STAN
language via an R interface. Two popular interfaces that greatly
simplify the specification of Bayesian models are
[brms](https://paul-buerkner.github.io/brms/) and
[rstanarm](https://mc-stan.org/rstanarm/). We will exclusively focus
on the former as it is far more flexible.

| Family            | Parameter                            | brms                            | rstanarm                    |
|-------------------|--------------------------------------|---------------------------------|-----------------------------|
| Gaussian          | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | Sigma                                | `student_t(3,0,mad(y))`         | `exponential(1/sd(y))`      |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |
| Poisson           | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |
| Negative binomial | Intercept                            | `student_t(3,median(y),mad(y))` | `normal(mean(y),2.5*sd(y))` |
|                   | 'Population effects' (slopes, betas) | flat, improper priors           | `normal(0,2.5*sd(y)/sd(x))` |
|                   | Shape                                | `gamma(0.01, 0.01)`             | `exponential(1/sd(y))`      |
|                   | 'Group-level effects'                | `student_t(3,0,mad(y))`         | `decov(1,1,1,1)`            |
|                   | Correlation on group-level effects   | `ljk_corr_cholesky(1)`          |                             |

: {.primary .bordered .sm .paramsTable}

Notes:

`brms`

https://github.com/paul-buerkner/brms/blob/c2b24475d727c8afd8bfc95947c18793b8ce2892/R/priors.R

1. In the above, for non-Gaussian families, `y` is first transformed
  according to the family link.  If the family link is `log`, then 0.1
  is first added to 0 values.
2. in `brms` the minimum standard deviation for the Intercept prior is
   `2.5`
3. in `brms` the minimum standard deviation for group-level priors is
   `10`.

`rstanarm`

http://mc-stan.org/rstanarm/articles/priors.html

1. in `rstanarm` priors on standard deviation and correlation
   associated with group-level effects are packaged up into a single
   prior (`decov` which is a decomposition of the variance and
   covariance matrix).

In my experience, I find that the above priors tend to be a little bit
too wide for many ecological applications and I often prefer to use
1.5 rather than 2.5 as the multiplier.

:::: {.callout-note}

In Bayesian models, centering of predictors offers huge numerical advantages.  So important is it to center that `brms` automatically centers any continuous predictors for you.  However, since the user has not necessarily centered the predictors, the user might misinterpret the outputs from a `brms` model.  Consequently, when fitting a model, `brms` also generates y-intercept values that are consistent with un-centered values and these are the estimates returned to the user.

Nevertheless, I would recommend that you should always explicitly
center continuous predictors to provide more meaningful
interpretations of the y-intercept. I would also highly recommend
standardising continuous predictors - this will not only help speed up
and stabilise the model, it will simplify the spefication of priors -
see the specific examples later in this tutorial.

::::

Based on the above, for our fabricated data, lets assign the following
priors:

- $\beta_0$: Normal prior centred at `r round(median(dat$y),2)` with a
  variance of `r round(mad(dat$y),2)`
  - mean: 
    ```{r}
    #| prior_y_mean_1a
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> median() |> round(2)
    ```
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    dat$y |> mad() |> round(2)
    ```
- $\beta_1$: _t_ distribution (3 degrees of freedom) prior centred at
  0 with a variance of `r round(mad(dat$y)/mad(dat$x),2)`
  - mean: since effects are differences and we may not want to
    pre-determine the direction (polarity) of any trends, it is
    typical to define effects priors with means of 0
  - variance: 
    ```{r}
    #| prior_y_var1_1a
    #| echo: true
    #| eval: true
    #| cache: false
    mad(dat$y) / mad(dat$x) |> round(2)
    ```
- $\sigma$: (half) _t_ distribution (3 degrees of freedom) centred at
  0 with a variance of `r round(mad(dat$y),2)`
  - variance: 
    ```{r}
    #| prior_y_var_1a
    #| echo: true
    #| eval: true
    #| cache: false
    ```

Note, again, when fitting models through either `rstanarm` or `brms`,
the priors assume that the predictor(s) have been centred and are to
be applied on the link scale. In this case the link scale is an
identity.



Similar logic can be applied for models that employ different
distributions. In the following sections, we will define numerous sets
of data (each of which represents different major forms of ecological
data) and see how we can set appropriate priors in each case. In
working through these example, it is worth reflecting on how much
simpler prior specification is if we use standardised predictors.


# Example data

This tutorial will blend theoretical discussions with actual
calculations and model fits. I believe that by bridging the divide
between theory and application, we all gain better understanding. The
applied components of this tutorial will be motivated by numerous
fabricated data sets. The advantage of simulated data over real data
is that with simulated data, we know the 'truth' and can therefore
gauge the accuracy of estimates.

The motivating examples are:

- **Example 1** - simulated samples drawn from a Gaussian (normal)
  distribution reminiscent of data collected on measurements (such as
  body mass)
- **Example 2** - simulated Gaussian samples drawn three different
  populations representing three different treatment levels (e.g. body
  masses of three different species)
- **Example 3** - simulated samples drawn from a Poisson distribution
  reminiscent of count data (such as number of individuals of a
  species within quadrats)
- **Example 4** - simulated samples drawn from a Negative Binomial
  distribution reminiscent of over-dispersed count data (such as
  number of individuals of a species that tends to aggregate in
  groups)
- **Example 5** - simulated samples drawn from a Bernoulli (binomial
  with $n = 1$) distribution reminiscent of binary data (such as the
  presence/absence of a species within sites)
- **Example 6** - simulated samples drawn from a Binomial distribution
  reminiscent of proportional data (such as counts of a particular
  taxa out of a total number of individuals)


# References

